---
title: "HPV Integration Classification Report"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 3
    code_folding: hide
    df_print: paged
params:
  # ── Required inputs ────────────────────────────────────────────────────────
  # ── Optional inputs (set to "" to skip) ────────────────────────────────────
  # ── Thresholds ─────────────────────────────────────────────────────────────
  # ── Output ─────────────────────────────────────────────────────────────────
  searchhpv_txt:        "data/searchHPV_breakpoints.txt"
  aa_profiles:          "data/amplicon_classification_profiles.tsv"
  purity_tsv:           "data/tumor_purity.tsv"     # columns: sample_id, tumor_purity
  coverage_tsv:         "data/viral_coverage.tsv"   # columns: sample_id, observed_viral_cov, expected_viral_cov
  min_sr_junc:          3      # minimum srNum to retain a breakpoint
  high_conf_only:       FALSE  # TRUE: restrict to High Confidence Site rows only
  vaf_purity_ratio_lo:  0.5    # VAF/purity below this → subclonal
  excess_coverage_fold: 1.5    # obs/expected viral cov above this → episomal modifier
  site_bin_bp:          200    # window (bp) for collapsing nearby breakpoints into one site
  internal_hpv_junc_min: 2     # min intra-HPV split reads to flag episomal rearrangement
  out_dir:              "results"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 9,
  fig.height = 5
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(ggplot2)
  library(knitr)
  library(kableExtra)
})

dir.create(params$out_dir,                        showWarnings = FALSE)
dir.create(file.path(params$out_dir, "plots"),    showWarnings = FALSE)

# ── Classification labels (8 categories) ──────────────────────────────────
# Drawn directly from the canonical classification table:
#   1. Episomal (non-integrated)
#   2. Episomal rearranged
#   3. Integrated clonal, single site
#   4. Integrated clonal, concatemeric single locus
#   5. Integrated clonal, multiple sites
#   6. Integrated subclonal
#   7. Integrated + episomal (mixed)
#   8. Human-HPV chimeric episome

class_palette <- c(
  "Episomal (non-integrated)"                = "#1D9E75",
  "Episomal rearranged"                      = "#3EC9A7",
  "Integrated clonal, single site"           = "#534AB7",
  "Integrated clonal, concatemeric"          = "#7F77DD",
  "Integrated clonal, multiple sites"        = "#3A2F8F",
  "Integrated subclonal"                     = "#D85A30",
  "Integrated + episomal (mixed)"            = "#BA7517",
  "Human-HPV chimeric episome"               = "#D4537E",
  "Unclassified"                             = "#888780"
)

# ── Assay flags: categories that CANNOT be definitively called from
#    short-read WGS alone (require long-read / optical mapping) ─────────────
long_read_required <- c(
  "Integrated clonal, concatemeric",
  "Integrated + episomal (mixed)",
  "Human-HPV chimeric episome"
)
```

------------------------------------------------------------------------

# Input parameters

All thresholds and file paths used for this run. Edit via `params:` in
the YAML header, or pass at render time:

``` r
rmarkdown::render("hpv_integration_classifier.Rmd",
                  params = list(searchhpv_txt = "my_file.txt", ...))
```

```{r show-params, echo=FALSE}
tibble(
  Parameter = names(unlist(params)),
  Value     = as.character(unlist(params))
) %>%
  kable(align = "ll") %>%
  kable_styling(bootstrap_options = c("striped", "condensed", "hover"),
                full_width = FALSE)
```

------------------------------------------------------------------------

# Classification scheme

The pipeline resolves each sample into one of **8 canonical
categories**, following the evidence hierarchy below. Categories marked
⚠️ cannot be definitively called from short-read WGS alone and require
long-read or optical mapping for confirmation.

| \# | Category | Key discriminator | Long-read required? |
|----|----|----|----|
| 1 | Episomal (non-integrated) | No human-viral junctions; uniform viral coverage; circular HPV-HPV reconstruction | No |
| 2 | Episomal rearranged | No human-viral junctions; intra-HPV junctions at non-canonical coordinates | No |
| 3 | Integrated clonal, single site | Single reciprocal junction pair; VAF ≈ purity; linear AA amplicon | No |
| 4 | Integrated clonal, concatemeric | Single locus; internal HPV-HPV head-to-tail junctions; no circularization | ⚠️ Yes |
| 5 | Integrated clonal, multiple sites | ≥2 reciprocal junction pairs at distinct loci; all at clonal VAF | No |
| 6 | Integrated subclonal | Junction VAF(s) below purity-adjusted expectation | No |
| 7 | Integrated + episomal (mixed) | Human-viral junctions present AND excess viral coverage (episomal pool) | ⚠️ Yes |
| 8 | Human-HPV chimeric episome | Human-viral junctions with non-contiguous/trans-chromosomal human flanks; circular AA reconstruction with human content | ⚠️ Yes |

------------------------------------------------------------------------

# 1 Load & parse data {.tabset}

## searchHPV breakpoints

```{r load-searchhpv}
shpv_raw <- read_tsv(params$searchhpv_txt, show_col_types = FALSE)

# ── Column mapping ─────────────────────────────────────────────────────────
# contigName                → contig_name
# orientation               → orientation   (hv / vh)
# genomeInsertionChromosome → chr_human
# genomeInsertionPoint      → pos_human
# hpvInsertionPoint         → pos_hpv
# SiteConfidence            → site_conf
# ContigConfidence          → contig_conf
# srNum                     → sr_num
# Sample                    → sample_full   (e.g. HPV02-RS-Pool_1_23__HPV16)
# ID1                       → junction_id
# Breakpoint_duplicate      → bp_duplicate

shpv <- shpv_raw %>%
  rename(
    contig_name  = contigName,
    chr_human    = genomeInsertionChromosome,
    pos_human    = genomeInsertionPoint,
    pos_hpv      = hpvInsertionPoint,
    site_conf    = SiteConfidence,
    contig_conf  = ContigConfidence,
    sr_num       = srNum,
    sample_full  = Sample,
    junction_id  = ID1,
    bp_duplicate = Breakpoint_duplicate
  ) %>%
  mutate(
    sample_id = str_remove(sample_full, "__HPV\\d+.*$"),
    hpv_type  = str_extract(sample_full, "HPV\\d+$")
  ) %>%
  filter(!is.na(sr_num), sr_num >= params$min_sr_junc)

if (params$high_conf_only) {
  shpv <- filter(shpv, site_conf == "High Confidence Site")
}

cat(sprintf("Loaded %d breakpoints across %d samples (srNum >= %d).\n",
            nrow(shpv), n_distinct(shpv$sample_id), params$min_sr_junc))
```

```{r preview-shpv, echo=FALSE}
head(shpv, 10) %>%
  select(sample_id, hpv_type, chr_human, pos_human, pos_hpv,
         orientation, sr_num, site_conf, contig_conf) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "condensed", "hover", "responsive"),
                full_width = TRUE) %>%
  scroll_box(height = "300px")
```

## AmpliconArchitect profiles

```{r load-aa}
# Required columns: sample_id, amplicon_id, amplicon_classification
# AA values used by this pipeline:
#   "Cyclic"               → circular reconstruction (episomal or chimeric ecDNA)
#   "Linear"               → chromosomal integration (no circularization)
#   "BFB"                  → breakage-fusion-bridge (treated as chromosomal)
#   "Heavily-rearranged"   → chromosomal
#   "No-amplification"     → no focal amplicon
#
# For chimeric ecDNA detection, the pipeline also looks for human sequence
# content within cyclic amplicons (aa_has_human_content column, if present).

aa <- read_tsv(params$aa_profiles, show_col_types = FALSE)

# Flag whether any cyclic amplicon per sample contains human sequence
# If your AA output has a column indicating human content (e.g. "human_content"
# or "contains_human"), rename it to aa_human_in_cyclic below.
# Otherwise this defaults to NA and chimeric ecDNA detection relies on
# junction coordinate dispersion alone.
if (!"aa_human_in_cyclic" %in% names(aa)) {
  aa <- aa %>%
    group_by(sample_id) %>%
    mutate(aa_human_in_cyclic = NA) %>%
    ungroup()
}

cat(sprintf("Loaded %d amplicon entries across %d samples.\n",
            nrow(aa), n_distinct(aa$sample_id)))
```

```{r preview-aa, echo=FALSE}
head(aa, 10) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "condensed", "hover"),
                full_width = FALSE)
```

## Tumor purity (optional)

```{r load-purity}
purity <- NULL
if (!is.null(params$purity_tsv) && nchar(params$purity_tsv) > 0 &&
    file.exists(params$purity_tsv)) {
  purity <- read_tsv(params$purity_tsv, show_col_types = FALSE)
  cat(sprintf("Loaded purity for %d samples.\n", nrow(purity)))
} else {
  cat("No purity file provided — VAF concordance (subclonal flag) will be 'unknown'.\n")
}
```

## Viral coverage (optional)

```{r load-coverage}
coverage <- NULL
if (!is.null(params$coverage_tsv) && nchar(params$coverage_tsv) > 0 &&
    file.exists(params$coverage_tsv)) {
  coverage <- read_tsv(params$coverage_tsv, show_col_types = FALSE)
  # Expected columns: sample_id, observed_viral_cov, expected_viral_cov
  # observed_viral_cov: mean depth across the viral genome
  # expected_viral_cov: depth predicted from junction VAF alone
  #   (i.e. junction_VAF * mean_host_depth)
  # Excess = episomal pool contribution
  cat(sprintf("Loaded coverage data for %d samples.\n", nrow(coverage)))
} else {
  cat("No coverage file provided — episomal modifier (Q5) will be 'unknown'.\n")
}
```

------------------------------------------------------------------------

# 2 Per-sample breakpoint summaries

Breakpoints within **`r params$site_bin_bp` bp** on the same chromosome
are collapsed into one integration site. A reciprocal pair requires at
least one `hv` and one `vh` orientation breakpoint per site.

```{r junc-summary}
junc_summary <- shpv %>%
  mutate(
    site_bin = paste(chr_human,
                     floor(pos_human / params$site_bin_bp) * params$site_bin_bp,
                     sep = ":")
  ) %>%
  group_by(sample_id) %>%
  summarise(
    n_breakpoints      = n(),
    n_sites            = n_distinct(site_bin),
    n_hv               = sum(orientation == "hv"),
    n_vh               = sum(orientation == "vh"),
    has_reciprocal     = n_hv > 0 & n_vh > 0,
    hpv_type           = first(hpv_type),
    high_conf_n        = sum(site_conf == "High Confidence Site"),
    max_sr             = max(sr_num, na.rm = TRUE),
    mean_sr            = round(mean(sr_num, na.rm = TRUE), 1),
    # Spread of human chromosomal coordinates — large spread or multi-chr
    # suggests chimeric ecDNA (non-contiguous human flanks)
    n_distinct_chr     = n_distinct(chr_human),
    chr_coord_range_mb = if (n_distinct(chr_human) == 1)
                           round((max(pos_human) - min(pos_human)) / 1e6, 2)
                         else NA_real_,
    .groups = "drop"
  )

junc_summary %>%
  kable(col.names = c("Sample", "Breakpoints", "Sites", "hv", "vh",
                      "Reciprocal?", "HPV type", "High-conf n",
                      "Max srNum", "Mean srNum",
                      "Distinct chr", "Coord range (Mb)")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed", "hover", "responsive"),
                full_width = TRUE) %>%
  scroll_box(height = "350px")
```

------------------------------------------------------------------------

# 3 Decision-tree functions

```{r decision-tree-functions}

# ── Q1a: Human-viral junctions present? ───────────────────────────────────
q1_junctions <- function(sid) {
  s <- filter(junc_summary, sample_id == sid)
  if (nrow(s) == 0) return(list(result = FALSE, n_bp = 0L, n_sites = 0L))
  list(result = TRUE, n_bp = s$n_breakpoints, n_sites = s$n_sites)
}

# ── Q1b: Intra-HPV (viral-viral) junctions at non-canonical coordinates?
#  Flags episomal rearrangement when no human-viral junctions exist.
#  Requires a column "intra_hpv_sr" in shpv (split reads where both
#  coordinates map within the HPV genome at non-adjacent positions).
#  If absent, this check is skipped and the sample stays "Episomal (non-integrated)".
q1b_intra_hpv <- function(sid) {
  if (!"intra_hpv_sr" %in% names(shpv)) return(list(result = FALSE, n = 0L))
  n <- shpv %>%
    filter(sample_id == sid, !is.na(intra_hpv_sr)) %>%
    pull(intra_hpv_sr) %>%
    sum(na.rm = TRUE)
  list(result = n >= params$internal_hpv_junc_min, n = as.integer(n))
}

# ── Q2: VAF concordant with tumor purity? ─────────────────────────────────
# VAF proxy = max(srNum) / 100  (replace with a true VAF column if available)
q2_concordance <- function(sid) {
  if (is.null(purity)) return(list(result = "unknown", ratio = NA_real_))
  pu <- filter(purity, sample_id == sid)
  if (nrow(pu) == 0 || is.na(pu$tumor_purity[1])) {
    return(list(result = "unknown", ratio = NA_real_))
  }
  js        <- filter(junc_summary, sample_id == sid)
  vaf_proxy <- js$max_sr[1] / 100
  ratio     <- vaf_proxy / pu$tumor_purity[1]
  list(
    result = if (ratio < params$vaf_purity_ratio_lo) "subclonal" else "concordant",
    ratio  = round(ratio, 3)
  )
}

# ── Q3: AA context — circular (ecDNA) or linear (chromosomal)? ────────────
q3_context <- function(sid) {
  sa <- filter(aa, sample_id == sid)
  if (nrow(sa) == 0) return(list(
    result         = "chromosomal",
    aa_class       = "not_in_AA",
    has_cyclic     = FALSE,
    has_human_cont = FALSE
  ))
  classes       <- sa$amplicon_classification
  has_cyclic    <- any(str_detect(classes, regex("cyclic", ignore_case = TRUE)))
  has_human_cont <- any(sa$aa_human_in_cyclic %in% TRUE, na.rm = TRUE)
  list(
    result         = if (has_cyclic) "cyclic" else "chromosomal",
    aa_class       = paste(sort(unique(classes)), collapse = ";"),
    has_cyclic     = has_cyclic,
    has_human_cont = has_human_cont
  )
}

# ── Q4: Internal HPV-HPV head-to-tail junctions? (concatemeric) ───────────
# Requires column "internal_hpv_hpv_junctions" (logical or count) in AA output
# or a dedicated column in shpv. Falls back to FALSE if absent.
q4_concatemeric <- function(sid) {
  if ("internal_hpv_hpv_junctions" %in% names(aa)) {
    sa  <- filter(aa, sample_id == sid)
    val <- any(sa$internal_hpv_hpv_junctions %in% TRUE, na.rm = TRUE)
    return(list(result = val))
  }
  list(result = FALSE)
}

# ── Q5: Single junction pair or multiple? ─────────────────────────────────
q5_multiplicity <- function(sid) {
  n <- filter(junc_summary, sample_id == sid)$n_sites[1]
  list(result = if (is.na(n) || n <= 1) "single" else "multiple", n_sites = n)
}

# ── Q6: Excess viral coverage (episomal modifier)? ────────────────────────
q6_excess_cov <- function(sid) {
  if (is.null(coverage)) return(list(result = "unknown", cov_ratio = NA_real_))
  sc <- filter(coverage, sample_id == sid)
  if (nrow(sc) == 0) return(list(result = "unknown", cov_ratio = NA_real_))
  ratio <- sc$observed_viral_cov[1] / sc$expected_viral_cov[1]
  list(
    result    = if (!is.na(ratio) && ratio > params$excess_coverage_fold) "yes" else "no",
    cov_ratio = round(ratio, 3)
  )
}

# ── Q7: Chimeric ecDNA? (non-contiguous / trans-chromosomal human flanks) ─
# Triggered when AA cyclic amplicon contains human content OR when junction
# human coordinates span multiple chromosomes / are very far apart.
q7_chimeric_ecdna <- function(sid, q3) {
  js <- filter(junc_summary, sample_id == sid)
  # Strong evidence: AA reports human content in cyclic amplicon
  if (isTRUE(q3$has_human_cont)) return(TRUE)
  # Supportive evidence: junctions span multiple chromosomes
  if (!is.na(js$n_distinct_chr) && js$n_distinct_chr > 1) return(TRUE)
  # Supportive evidence: single chromosome but flanks > 10 Mb apart
  # (inconsistent with a simple chromosomal insertion)
  if (!is.na(js$chr_coord_range_mb) && js$chr_coord_range_mb > 10) return(TRUE)
  FALSE
}
```

------------------------------------------------------------------------

# 4 Classify all samples

```{r classify}
classify_sample <- function(sid) {

  hpv_t <- filter(junc_summary, sample_id == sid)$hpv_type[1]
  if (is.na(hpv_t)) hpv_t <- shpv %>% filter(sample_id == sid) %>% pull(hpv_type) %>% first()

  q1  <- q1_junctions(sid)
  q1b <- q1b_intra_hpv(sid)

  # ── No human-viral junctions ──────────────────────────────────────────────
  if (!q1$result) {
    cat_label <- if (q1b$result) "Episomal rearranged" else "Episomal (non-integrated)"
    return(tibble(
      sample_id            = sid,
      classification       = cat_label,
      canonical_name       = cat_label,
      assay_note           = "Standard WGS sufficient",
      long_read_required   = FALSE,
      q1_has_junctions     = FALSE,
      q1_n_breakpoints     = 0L,
      q1_n_sites           = 0L,
      q1b_intra_hpv_sr     = as.integer(q1b$n),
      q2_concordance       = NA_character_,
      q2_vaf_ratio         = NA_real_,
      q3_aa_class          = NA_character_,
      q3_aa_cyclic         = NA,
      q3_aa_human_content  = NA,
      q4_concatemeric      = NA,
      q5_multiplicity      = NA_character_,
      q5_n_sites           = NA_integer_,
      q6_excess_cov        = NA_character_,
      q6_cov_ratio         = NA_real_,
      hpv_type             = hpv_t
    ))
  }

  # ── Human-viral junctions present — check VAF ─────────────────────────────
  q2 <- q2_concordance(sid)
  if (q2$result == "subclonal") {
    return(tibble(
      sample_id            = sid,
      classification       = "Integrated subclonal",
      canonical_name       = "Subclonal integration (any form)",
      assay_note           = "Deep coverage required (>200x for CCF ~0.3)",
      long_read_required   = FALSE,
      q1_has_junctions     = TRUE,
      q1_n_breakpoints     = as.integer(q1$n_bp),
      q1_n_sites           = as.integer(q1$n_sites),
      q1b_intra_hpv_sr     = as.integer(q1b$n),
      q2_concordance       = "subclonal",
      q2_vaf_ratio         = q2$ratio,
      q3_aa_class          = NA_character_,
      q3_aa_cyclic         = NA,
      q3_aa_human_content  = NA,
      q4_concatemeric      = NA,
      q5_multiplicity      = NA_character_,
      q5_n_sites           = NA_integer_,
      q6_excess_cov        = NA_character_,
      q6_cov_ratio         = NA_real_,
      hpv_type             = hpv_t
    ))
  }

  # ── Clonal junctions — check AA context ───────────────────────────────────
  q3 <- q3_context(sid)

  if (q3$result == "cyclic") {
    # Cyclic AA: chimeric ecDNA or pure episomal — distinguish by human content
    is_chimeric <- q7_chimeric_ecdna(sid, q3)
    if (is_chimeric) {
      return(tibble(
        sample_id            = sid,
        classification       = "Human-HPV chimeric episome",
        canonical_name       = "Heterocatemer / virus-host ecDNA",
        assay_note           = "DEFINITIVE CALL REQUIRES long-read + optical mapping",
        long_read_required   = TRUE,
        q1_has_junctions     = TRUE,
        q1_n_breakpoints     = as.integer(q1$n_bp),
        q1_n_sites           = as.integer(q1$n_sites),
        q1b_intra_hpv_sr     = as.integer(q1b$n),
        q2_concordance       = q2$result,
        q2_vaf_ratio         = q2$ratio,
        q3_aa_class          = q3$aa_class,
        q3_aa_cyclic         = TRUE,
        q3_aa_human_content  = q3$has_human_cont,
        q4_concatemeric      = NA,
        q5_multiplicity      = NA_character_,
        q5_n_sites           = NA_integer_,
        q6_excess_cov        = NA_character_,
        q6_cov_ratio         = NA_real_,
        hpv_type             = hpv_t
      ))
    }
  }

  # ── Chromosomal context: check for concatemeric integration ───────────────
  q4 <- q4_concatemeric(sid)

  if (q4$result) {
    return(tibble(
      sample_id            = sid,
      classification       = "Integrated clonal, concatemeric",
      canonical_name       = "Type 2 integration (McBride & Warburton 2017)",
      assay_note           = "DEFINITIVE CALL REQUIRES long-read or optical mapping",
      long_read_required   = TRUE,
      q1_has_junctions     = TRUE,
      q1_n_breakpoints     = as.integer(q1$n_bp),
      q1_n_sites           = as.integer(q1$n_sites),
      q1b_intra_hpv_sr     = as.integer(q1b$n),
      q2_concordance       = q2$result,
      q2_vaf_ratio         = q2$ratio,
      q3_aa_class          = q3$aa_class,
      q3_aa_cyclic         = q3$has_cyclic,
      q3_aa_human_content  = q3$has_human_cont,
      q4_concatemeric      = TRUE,
      q5_multiplicity      = NA_character_,
      q5_n_sites           = NA_integer_,
      q6_excess_cov        = NA_character_,
      q6_cov_ratio         = NA_real_,
      hpv_type             = hpv_t
    ))
  }

  # ── Single vs multi-site ───────────────────────────────────────────────────
  q5  <- q5_multiplicity(sid)
  q6  <- q6_excess_cov(sid)

  has_episomal_modifier <- !is.na(q6$result) && q6$result == "yes"

  if (has_episomal_modifier) {
    return(tibble(
      sample_id            = sid,
      classification       = "Integrated + episomal (mixed)",
      canonical_name       = "Mixed integrated/episomal",
      assay_note           = "Distinction from chimeric ecDNA requires long-read or optical mapping",
      long_read_required   = TRUE,
      q1_has_junctions     = TRUE,
      q1_n_breakpoints     = as.integer(q1$n_bp),
      q1_n_sites           = as.integer(q1$n_sites),
      q1b_intra_hpv_sr     = as.integer(q1b$n),
      q2_concordance       = q2$result,
      q2_vaf_ratio         = q2$ratio,
      q3_aa_class          = q3$aa_class,
      q3_aa_cyclic         = q3$has_cyclic,
      q3_aa_human_content  = q3$has_human_cont,
      q4_concatemeric      = FALSE,
      q5_multiplicity      = q5$result,
      q5_n_sites           = as.integer(q5$n_sites),
      q6_excess_cov        = "yes",
      q6_cov_ratio         = q6$cov_ratio,
      hpv_type             = hpv_t
    ))
  }

  # ── Simple chromosomal integration: single vs multi-site ──────────────────
  final_class <- if (q5$result == "single") {
    "Integrated clonal, single site"
  } else {
    "Integrated clonal, multiple sites"
  }
  canonical <- if (q5$result == "single") {
    "Type 1 integration (McBride & Warburton 2017)"
  } else {
    "Multi-locus Type 1 integration"
  }

  tibble(
    sample_id            = sid,
    classification       = final_class,
    canonical_name       = canonical,
    assay_note           = "Standard WGS sufficient; long-read aids breakpoint resolution",
    long_read_required   = FALSE,
    q1_has_junctions     = TRUE,
    q1_n_breakpoints     = as.integer(q1$n_bp),
    q1_n_sites           = as.integer(q1$n_sites),
    q1b_intra_hpv_sr     = as.integer(q1b$n),
    q2_concordance       = q2$result,
    q2_vaf_ratio         = q2$ratio,
    q3_aa_class          = q3$aa_class,
    q3_aa_cyclic         = q3$has_cyclic,
    q3_aa_human_content  = q3$has_human_cont,
    q4_concatemeric      = FALSE,
    q5_multiplicity      = q5$result,
    q5_n_sites           = as.integer(q5$n_sites),
    q6_excess_cov        = q6$result,
    q6_cov_ratio         = q6$cov_ratio,
    hpv_type             = hpv_t
  )
}

all_samples <- union(unique(shpv$sample_id), unique(aa$sample_id))
results     <- map_dfr(all_samples, classify_sample)

write_tsv(results, file.path(params$out_dir, "summary_table.tsv"))
cat(sprintf("Classified %d samples. Written to %s/summary_table.tsv\n",
            nrow(results), params$out_dir))
```

------------------------------------------------------------------------

# 5 Results {.tabset}

## Classification summary

```{r results-summary}
results %>%
  count(classification, canonical_name, assay_note, long_read_required,
        name = "Samples") %>%
  arrange(desc(Samples)) %>%
  mutate(long_read_required = if_else(long_read_required, "⚠️ Yes", "No")) %>%
  kable(col.names = c("Classification", "Canonical name", "Assay note",
                      "Long-read required?", "Samples"),
        align = "llllr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) %>%
  column_spec(4, bold = TRUE,
              color = if_else(
                results %>%
                  count(classification, long_read_required) %>%
                  arrange(desc(n)) %>%
                  pull(long_read_required),
                "darkorange", "black"))
```

## Long-read flagged samples

Samples in categories that **cannot be definitively resolved by
short-read WGS**:

```{r results-longread}
lr_df <- filter(results, long_read_required)
if (nrow(lr_df) > 0) {
  lr_df %>%
    select(sample_id, hpv_type, classification, assay_note,
           q1_n_breakpoints, q3_aa_class) %>%
    kable(col.names = c("Sample", "HPV type", "Classification",
                        "Assay note", "Breakpoints", "AA class")) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = TRUE) %>%
    row_spec(0, background = "#fff3cd")
} else {
  cat("No samples flagged as requiring long-read confirmation.")
}
```

## searchHPV breakpoints

```{r results-junctions}
results %>%
  filter(q1_has_junctions) %>%
  select(sample_id, hpv_type, q1_n_breakpoints, q1_n_sites,
         q1b_intra_hpv_sr, classification) %>%
  arrange(desc(q1_n_breakpoints)) %>%
  kable(col.names = c("Sample", "HPV type", "Breakpoints", "Sites",
                      "Intra-HPV sr", "Classification")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = TRUE) %>%
  scroll_box(height = "400px")
```

## VAF / purity concordance

```{r results-vaf}
vdf <- filter(results, !is.na(q2_vaf_ratio))
if (nrow(vdf) > 0) {
  vdf %>%
    select(sample_id, q2_concordance, q2_vaf_ratio, classification) %>%
    kable(col.names = c("Sample", "Concordance", "VAF/purity ratio", "Classification"),
          digits = 3) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE)
} else {
  cat("No purity data provided — Q2 skipped.")
}
```

## AmpliconArchitect context

```{r results-aa}
results %>%
  filter(!is.na(q3_aa_class)) %>%
  count(q3_aa_class, q3_aa_cyclic, q3_aa_human_content, classification) %>%
  mutate(across(c(q3_aa_cyclic, q3_aa_human_content), as.character)) %>%
  kable(col.names = c("AA class(es)", "Cyclic?", "Human content in cyclic?",
                      "Classification", "N"),
        align = "llllr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

## Multiplicity & coverage

```{r results-multiplicity}
results %>%
  filter(!is.na(q5_multiplicity)) %>%
  select(sample_id, hpv_type, q5_multiplicity, q5_n_sites,
         q6_excess_cov, q6_cov_ratio, classification) %>%
  kable(col.names = c("Sample", "HPV type", "Multiplicity", "N sites",
                      "Excess cov?", "Cov ratio", "Classification"),
        digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = TRUE) %>%
  scroll_box(height = "400px")
```

## Full results table

```{r results-full}
results %>%
  kable(digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = TRUE) %>%
  scroll_box(height = "500px")
```

------------------------------------------------------------------------

# 6 Plots {.tabset}

## Classification frequency

```{r plot-bar, fig.height=5}
p_bar <- results %>%
  count(classification, long_read_required) %>%
  mutate(
    classification = fct_reorder(classification, n),
    lr_label       = if_else(long_read_required, "⚠ Long-read required", "Short-read sufficient")
  ) %>%
  ggplot(aes(x = n, y = classification, fill = classification)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = n), hjust = -0.3, size = 3.5) +
  geom_text(aes(x = 0.1, label = if_else(long_read_required, "⚠", "")),
            hjust = 0, size = 4, colour = "darkorange") +
  scale_fill_manual(values = class_palette) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "HPV integration classification (8-category scheme)",
       subtitle = "⚠ = long-read / optical mapping required for definitive call",
       x = "Samples", y = NULL) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.y = element_blank())

ggsave(file.path(params$out_dir, "plots", "01_classification_bar.pdf"),
       p_bar, width = 9, height = 6)
p_bar
```

## Breakpoint counts

```{r plot-junc}
p_junc <- results %>%
  filter(!is.na(q1_n_breakpoints), q1_n_breakpoints > 0) %>%
  ggplot(aes(x = q1_n_breakpoints, fill = classification)) +
  geom_histogram(bins = 20, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = class_palette, name = NULL) +
  labs(title = "Breakpoints per sample (searchHPV)",
       x = paste0("Breakpoints (srNum >= ", params$min_sr_junc, ")"),
       y = "Samples") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom", legend.text = element_text(size = 8))

ggsave(file.path(params$out_dir, "plots", "02_junction_counts.pdf"),
       p_junc, width = 8, height = 5)
p_junc
```

## Site multiplicity by HPV type

```{r plot-multi}
p_multi <- results %>%
  filter(!is.na(q5_multiplicity)) %>%
  count(q5_multiplicity, hpv_type, classification) %>%
  ggplot(aes(x = hpv_type, y = n, fill = classification)) +
  geom_col(position = "stack", width = 0.6) +
  facet_wrap(~q5_multiplicity, labeller = label_both) +
  scale_fill_manual(values = class_palette, name = NULL) +
  labs(title = "Site multiplicity by HPV type",
       x = "HPV type", y = "Samples") +
  theme_minimal(base_size = 13) +
  theme(legend.position  = "bottom",
        legend.text      = element_text(size = 8),
        axis.text.x      = element_text(angle = 30, hjust = 1))

ggsave(file.path(params$out_dir, "plots", "03_site_multiplicity.pdf"),
       p_multi, width = 8, height = 5)
p_multi
```

## Confidence breakdown

```{r plot-conf, fig.height=4}
p_conf <- shpv %>%
  count(site_conf, contig_conf) %>%
  ggplot(aes(x = site_conf, y = n, fill = contig_conf)) +
  geom_col(position = "dodge", width = 0.6) +
  scale_fill_brewer(palette = "Set2", name = "Contig confidence") +
  labs(title = "searchHPV site & contig confidence breakdown",
       x = "Site confidence", y = "Breakpoints") +
  theme_minimal(base_size = 13)

ggsave(file.path(params$out_dir, "plots", "04_confidence_breakdown.pdf"),
       p_conf, width = 7, height = 5)
p_conf
```

## Long-read flag summary

```{r plot-longread, fig.height=4}
p_lr <- results %>%
  count(classification, long_read_required) %>%
  mutate(
    classification = fct_reorder(classification, n),
    assay          = if_else(long_read_required,
                             "Long-read / optical mapping required",
                             "Short-read WGS sufficient")
  ) %>%
  ggplot(aes(x = n, y = classification, fill = assay)) +
  geom_col(width = 0.65) +
  scale_fill_manual(values = c("Long-read / optical mapping required" = "#D85A30",
                               "Short-read WGS sufficient"           = "#534AB7"),
                    name = NULL) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Assay requirement by classification",
       x = "Samples", y = NULL) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.y = element_blank(),
        legend.position    = "bottom")

ggsave(file.path(params$out_dir, "plots", "05_assay_requirement.pdf"),
       p_lr, width = 9, height = 5)
p_lr
```

------------------------------------------------------------------------

# 7 Session info

```{r session-info, echo=FALSE}
sessionInfo()
```
