---
title: "HPV State Classification + Integration Locus Clustering — OPC (HN-keyed)"
subtitle: "searcHPV V2 + coverage z-scores; related-integration clustering by human locus"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 3
    code_folding: show
    df_print: paged
params:
  # ── Inputs ────────────────────────────────────────────────────────────────
  summary_xlsx:  "data/summ_OPC.v7.2.xlsx"                 # crosswalk: seqname <-> HN id
  searchhpv_txt: "data/2024-05-27_SearcHPV_OPC_V2.txt"     # OPC integration (V2 schema)
  zscore_csv:    "data/20241205_Zscores_stats_nofilter.csv" # OPC per-base coverage z-scores
  aa_tsv:        "data/merged_result_tables.tsv"            # AA (NOT OPC yet) - only used if use_aa
  # ── The toggle you asked for ──────────────────────────────────────────────
  use_aa:             false   # FALSE = run on searcHPV + coverage only (current OPC state)
  # ── Timepoint filter ──────────────────────────────────────────────────────
  timepoint_filter:   "Baseline"   # keep only this timepoint; set "" or null for all
  # ── Thresholds (tune here) ────────────────────────────────────────────────
  high_conf_site_only: false  # TRUE = keep only "High Confidence Site" searcHPV breakpoints
  site_bin_bp:        200     # window (bp) to collapse nearby breakpoints into one site
  cluster_dist_bp:    100000  # related-integration clustering: same HPV type + chr within this distance = one locus
  episomal_margin_bp: 200     # bp margin around the junction-spanned arc when looking for episomal coverage
  episomal_present_frac: 0.5  # fraction of outside-arc bases that must show coverage (z >= -z_cut) to call episomal HPV present
  z_cut:              1.5     # |z| >= this = coverage dip (negative) or gain (positive)
  uniform_frac:       0.05    # if < this fraction of bases deviate, coverage is "uniform"
  excess_gain_frac:   0.50    # if >= this fraction of bases gained, treat as episomal pool
  distant_flank_mb:   10      # human flanks farther apart than this (same chr) = chimeric signal
  focal_cn:           10      # AA max copy number above this = focal amplification
  overlap_window_bp:  200     # window (bp) to call a coverage/junction/AA overlap
  junc_cov_window_bp: 200     # window (bp) for junction-vs-coverage concordance check
  cov_block_merge_bp: 500     # merge same-type coverage segments within this gap into one block
  cov_gap_tol:        25      # bp: merge elevated runs separated by gaps <= this
  cov_smooth_bp:      75      # bp: rolling-mean window applied before segmenting (kills noise spikes); 1 = off
  cov_min_seg_len:    150     # bp: discard sustained segments shorter than this (noise)
  n_plot_samples:     220       # how many integrated samples to draw track PDFs for
  out_dir:            "results"   # TSV tables
  fig_dir:            "plots"     # per-sample plot PDFs go in this sub-directory
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6.5)
```

This version runs the **real OPC data** and uses the **HN identifier**
(the summary's `id` = `Patient_ID`\_`Timepoint`, e.g. `HN2186_Baseline`)
as the reference key throughout. The three sources are tied together by
the summary crosswalk:

-   **searcHPV** (`2024-05-27_SearcHPV_OPC_V2.txt`) — integration
    junctions. Its `Sample` is a sequencing-library name; it also
    carries `Patient_ID`, `Timepoint`, and a ready-made `id`.
-   **Coverage** (`20241205_Zscores_stats_nofilter.csv`) — per-base
    z-scores, keyed only by sequencing-library name (`file.name`).
-   **Summary** (`summ_OPC.v7.2.xlsx`) — maps every sequencing-library
    `Sample` to its HN `id`, `Patient_ID`, and `Timepoint`. This is the
    crosswalk that lets coverage be relabelled to HN ids.

> **HN id is per-timepoint.** Because integration and coverage differ
> between a patient's timepoints, the working key is the full `id` (HN +
> timepoint). The bare HN (`Patient_ID`) is also carried as a column so
> you can group across timepoints.

## The AA toggle

`params$use_aa` switches AmpliconArchitect on or off. OPC has no AA yet,
so the default is **`use_aa: false`** and the run uses searcHPV +
coverage only. When you have OPC AA, point `params$aa_tsv` at it and set
`use_aa: true` (you can also flip it in the RStudio *Knit with
Parameters* dialog).

**What AA off costs you.** AA is the only source for circular-molecule
evidence, so with it off the classifier cannot positively call
`Human-HPV chimeric episome` or the AA-driven
`Integrated + episomal (mixed)`. Those samples instead resolve to the
best junction/coverage call (usually multi-site integration), and the
`notes` column records that AA was unavailable. Everything else
(episomal vs rearranged, single vs multiple integration sites) is
unaffected.

------------------------------------------------------------------------

# 1 Packages and labels

```{r packages}
suppressPackageStartupMessages({
  library(tidyverse)
  library(readxl)
  library(knitr)
  library(kableExtra)
})

dir.create(params$out_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(params$fig_dir, showWarnings = FALSE, recursive = TRUE)

state_levels <- c(
  "Episomal only",
  "HPV episomal rearranged",
  "Human-HPV chimeric episome",
  "Integrated clonal, single site",
  "Integrated clonal, single site + episomal HPV",
  "Integrated clonal, multiple sites",
  "Integrated clonal, multiple sites + episomal HPV",
  "Integrated + episomal (mixed)",
  "Unclassified"
)
state_palette <- c(
  "Episomal only"                                    = "#1D9E75",
  "HPV episomal rearranged"                          = "#3EC9A7",
  "Human-HPV chimeric episome"                       = "#D4537E",
  "Integrated clonal, single site"                   = "#534AB7",
  "Integrated clonal, single site + episomal HPV"    = "#8C84E0",
  "Integrated clonal, multiple sites"                = "#3A2F8F",
  "Integrated clonal, multiple sites + episomal HPV" = "#6E63C9",
  "Integrated + episomal (mixed)"                    = "#BA7517",
  "Unclassified"                                     = "#888780"
)

cat(sprintf("AmpliconArchitect: %s\n", if (isTRUE(params$use_aa)) "ENABLED" else "DISABLED (searcHPV + coverage only)"))
```

------------------------------------------------------------------------

# 2 Crosswalk (HN id is the reference)

```{r crosswalk}
# NOTE: the sheet has TWO columns named "Sample" (col 1 = full sequencing name
# with _L00x, col 2 = without). readxl renames duplicates, so we grab the first
# by position - that is the one matching coverage file.name and searcHPV Sample.
summ_raw <- suppressMessages(read_excel(params$summary_xlsx, .name_repair = "unique"))
seqname_col <- names(summ_raw)[1]

summ <- summ_raw %>%
  transmute(seqname   = as.character(.data[[seqname_col]]),
            sample_id = as.character(id),          # HN_Timepoint = the reference key
            patient   = as.character(Patient_ID),  # bare HN
            timepoint = as.character(Timepoint),
            hpv_type  = as.character(HPV_realign))

# Timepoint filter (default Baseline). Cascades to coverage (via the crosswalk)
# and to searcHPV (filtered in the load chunk).
tp_filter <- params$timepoint_filter
if (!is.null(tp_filter) && nzchar(tp_filter)) {
  summ <- filter(summ, timepoint == tp_filter)
  cat(sprintf("Timepoint filter: keeping only '%s'.\n", tp_filter))
} else {
  cat("Timepoint filter: none (all timepoints).\n")
}

# seqname -> HN id, and the reverse (id is unique per seqname)
seq2id <- setNames(summ$sample_id, summ$seqname)
id2seq <- setNames(summ$seqname,   summ$sample_id)

meta <- summ %>% distinct(sample_id, patient, timepoint, hpv_type)

cat(sprintf("Crosswalk: %d sequencing libraries -> %d HN ids.\n",
            nrow(summ), n_distinct(summ$sample_id)))
head(summ, 5) %>% kable() %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

------------------------------------------------------------------------

# 3 Parsers

Each parser returns a per-sample feature table keyed by the HN
`sample_id`.

## 3.1 searcHPV (V2 schema)

The V2 file has no `orientation` or `srNum` columns, so reciprocal-pair
and read-count filters from earlier versions are dropped. Breakpoints
are filtered by `SiteConfidence` instead (optional, via
`high_conf_site_only`), and the HN id is taken from the file's own `id`
column when present, otherwise resolved through the crosswalk, otherwise
derived from the sample name (so synthetic tests still work).

```{r parse-searchpv}
parse_searchpv <- function(df,
                           high_conf_site_only = params$high_conf_site_only,
                           bin_bp = params$site_bin_bp) {
  empty <- tibble(sample_id = character(), hpv_type = character(),
                  n_breakpoints = integer(), n_sites = integer(),
                  n_distinct_chr = integer(), chr_coord_range_mb = double(),
                  flanks_noncontiguous = logical(), has_hv_junctions = logical())
  if (is.null(df) || nrow(df) == 0) return(empty)

  sid <- if ("id" %in% names(df)) as.character(df$id) else
           ifelse(as.character(df$Sample) %in% names(seq2id),
                  unname(seq2id[as.character(df$Sample)]),
                  str_remove(as.character(df$Sample), "__HPV\\d+.*$"))
  hpv <- if ("HPV_realign" %in% names(df)) as.character(df$HPV_realign) else
           str_extract(as.character(df$Sample), "HPV\\d+")
  conf <- if ("SiteConfidence" %in% names(df)) as.character(df$SiteConfidence) else NA_character_

  work <- tibble(
    sample_id = sid,
    hpv_type  = hpv,
    chr       = as.character(df$genomeInsertionChromosome),
    pos       = suppressWarnings(as.numeric(str_trim(as.character(df$genomeInsertionPoint)))),
    site_conf = conf
  )
  if (isTRUE(high_conf_site_only) && any(!is.na(work$site_conf)))
    work <- filter(work, str_detect(tolower(coalesce(site_conf, "")), "high"))
  if (nrow(work) == 0) return(empty)

  work %>%
    mutate(site_bin = paste(chr, floor(pos / bin_bp) * bin_bp, sep = ":")) %>%
    group_by(sample_id) %>%
    summarise(
      hpv_type           = first(na.omit(hpv_type)),
      n_breakpoints      = n(),
      n_sites            = n_distinct(site_bin),
      n_distinct_chr     = n_distinct(chr),
      chr_coord_range_mb = if (n_distinct(chr) == 1)
                             (max(pos, na.rm = TRUE) - min(pos, na.rm = TRUE)) / 1e6
                           else NA_real_,
      .groups = "drop"
    ) %>%
    mutate(
      flanks_noncontiguous = (n_distinct_chr > 1) |
                             (!is.na(chr_coord_range_mb) &
                                chr_coord_range_mb > params$distant_flank_mb),
      has_hv_junctions = n_breakpoints > 0
    )
}
```

## 3.2 Coverage z-scores

```{r parse-cov}
parse_cov <- function(df, z = params$z_cut,
                      uniform_frac = params$uniform_frac,
                      excess_frac  = params$excess_gain_frac) {
  empty <- tibble(seqname = character(), cov_frac_gain = double(),
                  cov_frac_dip = double(), cov_mean_z = double(),
                  cov_uniform = logical(), cov_n_gain_seg = integer(),
                  cov_n_dip_seg = integer(), cov_n_event_seg = integer(),
                  cov_excess = logical())
  if (is.null(df) || nrow(df) == 0) return(empty)

  id_col <- names(df)[1]                       # "file.name"
  ids    <- as.character(df[[id_col]])
  mat    <- as.matrix(df[, setdiff(names(df), id_col)])
  storage.mode(mat) <- "double"

  frac_gain <- rowMeans(mat >= z,  na.rm = TRUE)
  frac_dip  <- rowMeans(mat <= -z, na.rm = TRUE)
  mean_z    <- rowMeans(mat,        na.rm = TRUE)

  count_seg <- function(v, up) {
    flag <- if (up) (!is.na(v) & v >= z) else (!is.na(v) & v <= -z)
    if (!any(flag)) return(0L)
    sum(rle(flag)$values)
  }
  n_gain_seg <- apply(mat, 1, count_seg, up = TRUE)
  n_dip_seg  <- apply(mat, 1, count_seg, up = FALSE)

  tibble(
    seqname         = ids,
    cov_frac_gain   = frac_gain,
    cov_frac_dip    = frac_dip,
    cov_mean_z      = mean_z,
    cov_uniform     = (frac_gain + frac_dip) < uniform_frac,
    cov_n_gain_seg  = as.integer(n_gain_seg),
    cov_n_dip_seg   = as.integer(n_dip_seg),
    cov_n_event_seg = as.integer(n_gain_seg + n_dip_seg),
    cov_excess      = frac_gain >= excess_frac
  )
}
```

## 3.3 AmpliconArchitect (only read when `use_aa`)

```{r parse-aa}
parse_aa <- function(df, focal_cn = params$focal_cn) {
  empty <- tibble(sample_id = character(), aa_has_hpv = logical(),
                  aa_hpv_cyclic = logical(), aa_cyclic_has_human = logical(),
                  aa_max_cn = double(), aa_focal_amp = logical())
  if (is.null(df) || nrow(df) == 0) return(empty)

  sid <- as.character(df$`Sample name`)
  sid <- ifelse(sid %in% names(seq2id), unname(seq2id[sid]), sid)  # map to HN id if possible

  df %>%
    mutate(.sid = sid) %>%
    transmute(
      sample_id     = .sid,
      has_hpv_loc   = str_detect(tolower(coalesce(Location, "")), "hpv"),
      has_human_loc = str_detect(coalesce(Location, ""), "chr"),
      is_cyclic     = str_detect(tolower(coalesce(Classification, "")), "cyclic"),
      is_virus      = str_detect(tolower(coalesce(Classification, "")), "virus"),
      max_cn        = suppressWarnings(as.numeric(`Feature maximum copy number`))
    ) %>%
    group_by(sample_id) %>%
    summarise(
      aa_has_hpv          = any(has_hpv_loc | is_virus),
      aa_hpv_cyclic       = any(is_cyclic & has_hpv_loc),
      aa_cyclic_has_human = any(is_cyclic & has_hpv_loc & has_human_loc),
      aa_max_cn           = suppressWarnings(max(max_cn, na.rm = TRUE)),
      .groups = "drop"
    ) %>%
    mutate(aa_max_cn = ifelse(is.finite(aa_max_cn), aa_max_cn, NA_real_),
           aa_focal_amp = !is.na(aa_max_cn) & aa_max_cn >= focal_cn)
}
```

------------------------------------------------------------------------

# 4 Load real data

```{r load}
# searcHPV (apply the same timepoint filter when the column is present)
shpv_raw <- read_tsv(params$searchhpv_txt, show_col_types = FALSE)
if (!is.null(tp_filter) && nzchar(tp_filter) && "Timepoint" %in% names(shpv_raw))
  shpv_raw <- filter(shpv_raw, as.character(Timepoint) == tp_filter)
shpv_features <- parse_searchpv(shpv_raw)

# coverage (large file; read once, keep raw for plotting)
cov_raw       <- read_csv(params$zscore_csv, show_col_types = FALSE)
cov_features  <- parse_cov(cov_raw) %>%
  mutate(sample_id = unname(seq2id[seqname])) %>%   # seqname -> HN id
  filter(!is.na(sample_id)) %>%
  select(sample_id, everything(), -seqname)

# AA only if toggled on
aa_raw <- if (isTRUE(params$use_aa) && file.exists(params$aa_tsv))
  read_tsv(params$aa_tsv, show_col_types = FALSE) else NULL
aa_features <- if (isTRUE(params$use_aa)) parse_aa(aa_raw) else parse_aa(NULL)

cat(sprintf("searcHPV: %d integrated samples\ncoverage: %d samples (HN-mapped)\nAA: %s\n",
            nrow(shpv_features), nrow(cov_features),
            if (isTRUE(params$use_aa)) sprintf("%d samples", nrow(aa_features)) else "disabled"))
```

```{r unify}
build_unified <- function(shpv, aa, cov, meta) {
  ids <- unique(c(shpv$sample_id, aa$sample_id, cov$sample_id))
  tibble(sample_id = ids) %>%
    left_join(shpv, by = "sample_id") %>%
    left_join(aa,   by = "sample_id") %>%
    left_join(cov,  by = "sample_id") %>%
    left_join(meta, by = "sample_id", suffix = c("", "_meta")) %>%
    mutate(
      hpv_type         = coalesce(hpv_type, hpv_type_meta),
      has_hv_junctions = coalesce(has_hv_junctions, FALSE),
      n_breakpoints    = coalesce(n_breakpoints, 0L),
      n_sites          = coalesce(n_sites, 0L)
    ) %>%
    select(-any_of("hpv_type_meta"))
}

unified <- build_unified(shpv_features, aa_features, cov_features, meta)
cat(sprintf("Unified OPC table: %d samples.\n", nrow(unified)))
```

## 4.1 Positional helpers (coverage segments, junction/coverage concordance)

These operate on the shared HPV axis and are used by both the
concordance check (Section 6) and the plots (Section 7).

```{r positional-helpers}
# Centered rolling mean (NA-aware, no extra packages). Used to suppress single-
# base noise spikes before segmenting, so only sustained shifts are detected.
roll_mean <- function(x, w) {
  w <- as.integer(w)
  if (w <= 1L) return(x)
  if (w %% 2L == 0L) w <- w + 1L
  k <- w %/% 2L; n <- length(x)
  cs <- c(0, cumsum(ifelse(is.na(x), 0, x)))
  cn <- c(0, cumsum(as.integer(!is.na(x))))
  vapply(seq_len(n), function(i) {
    a <- max(1L, i - k); b <- min(n, i + k)
    cnt <- cn[b + 1L] - cn[a]
    if (cnt > 0L) (cs[b + 1L] - cs[a]) / cnt else NA_real_
  }, numeric(1))
}

seg_from_flag <- function(flag, gap_tol) {
  idx <- which(flag)
  if (length(idx) == 0) return(list())
  brk <- c(0L, which(diff(idx) > (gap_tol + 1L)), length(idx))
  lapply(seq_len(length(brk) - 1L), function(j) {
    grp <- idx[(brk[j] + 1L):brk[j + 1L]]; c(min(grp), max(grp))
  })
}
cov_segments <- function(zvec, cut = params$z_cut,
                         gap_tol = params$cov_gap_tol, min_len = params$cov_min_seg_len,
                         smooth = params$cov_smooth_bp) {
  z <- roll_mean(zvec, smooth)          # smooth first; detection runs on this
  out <- list()
  for (dir in c(1L, -1L)) {
    flag <- if (dir == 1L) (!is.na(z) & z >= cut) else (!is.na(z) & z <= -cut)
    for (m in seg_from_flag(flag, gap_tol)) {
      s <- m[1]; e <- m[2]
      if (e - s + 1L < min_len) next
      mz <- mean(z[s:e], na.rm = TRUE)
      if ((dir == 1L && mz >= cut) || (dir == -1L && mz <= -cut))
        out[[length(out) + 1L]] <- tibble(seg_start = s, seg_end = e,
                                          type = if (dir == 1L) "gain" else "dip")
    }
  }
  if (length(out) == 0) return(tibble(seg_start = integer(), seg_end = integer(), type = character()))
  bind_rows(out) %>% arrange(seg_start)
}

# Merge same-type segments separated by <= merge_gap into contiguous BLOCKS.
# Raw cov_segments are fragmented by sub-threshold wiggles inside one real block;
# block-merging gives the boundaries that should line up with junctions.
cov_blocks <- function(zvec, merge_gap = params$cov_block_merge_bp, ...) {
  segs <- cov_segments(zvec, ...)
  if (nrow(segs) == 0) return(segs)
  segs <- arrange(segs, type, seg_start)
  out <- list(); cur <- NULL
  for (i in seq_len(nrow(segs))) {
    s <- segs[i, ]
    if (is.null(cur)) { cur <- s; next }
    if (s$type == cur$type && (s$seg_start - cur$seg_end) <= merge_gap) {
      cur$seg_end <- max(cur$seg_end, s$seg_end)
    } else { out[[length(out) + 1L]] <- cur; cur <- s }
  }
  out[[length(out) + 1L]] <- cur
  bind_rows(out) %>% arrange(seg_start)
}

# Coverage vector for one HN id (NA-trimmed at the tail).
cov_vec <- function(sid) {
  sq <- unname(id2seq[sid])          # single-bracket: returns NA if id absent
  if (length(sq) == 0 || is.na(sq)) return(numeric())
  row <- cov_raw[as.character(cov_raw[[1]]) == sq, , drop = FALSE]
  if (nrow(row) == 0) return(numeric())
  v <- as.numeric(row[1, -1])
  last <- suppressWarnings(max(which(!is.na(v))))
  if (is.finite(last)) v[seq_len(last)] else v
}

# searcHPV breakpoints for one HN id: viral coordinate (hpv_pos) plus the human
# locus (chr, human_pos) and HPV type, used for the integration-locus clustering.
shpv_real <- shpv_raw %>%
  mutate(sample_id = if ("id" %in% names(.)) as.character(id) else
                       ifelse(as.character(Sample) %in% names(seq2id),
                              unname(seq2id[as.character(Sample)]),
                              str_remove(as.character(Sample), "__HPV\\d+.*$")),
         hpv_pos   = suppressWarnings(as.numeric(str_trim(as.character(hpvInsertionPoint)))),
         chr       = as.character(genomeInsertionChromosome),
         human_pos = suppressWarnings(as.numeric(str_trim(as.character(genomeInsertionPoint)))),
         hpv_type  = if ("HPV_realign" %in% names(.)) as.character(HPV_realign)
                     else str_extract(as.character(Sample), "HPV\\d+"))
junc_pos <- function(sid) shpv_real %>% filter(sample_id == sid, !is.na(hpv_pos)) %>% pull(hpv_pos)

# ── Related-integration clustering (human locus) ──────────────────────────────
# Within a sample, group breakpoints of the SAME HPV type on the SAME chromosome
# whose human positions are within cluster_dist_bp (single-linkage). Each group is
# one integration locus; two breakpoints in the same group are "related", separate
# groups are independent loci. Cross-chromosome = always separate.
breakpoint_clusters <- function(sid, dist = params$cluster_dist_bp) {
  d <- shpv_real %>% filter(sample_id == sid, !is.na(human_pos), !is.na(chr))
  if (nrow(d) == 0) return(tibble())
  d %>%
    arrange(hpv_type, chr, human_pos) %>%
    group_by(hpv_type, chr) %>%
    mutate(local_cl = cumsum(c(TRUE, diff(human_pos) > dist))) %>%
    group_by(hpv_type, chr, local_cl) %>%
    mutate(cluster_id = sprintf("%s:%.2fMb", chr[1], min(human_pos) / 1e6)) %>%
    ungroup()
}

# One row per integration locus for a sample.
integration_clusters <- function(sid, dist = params$cluster_dist_bp) {
  bp <- breakpoint_clusters(sid, dist)
  if (nrow(bp) == 0) return(tibble())
  bp %>%
    group_by(hpv_type, chr, cluster_id) %>%
    summarise(n_breakpoints = n(),
              human_start = min(human_pos), human_end = max(human_pos),
              span_bp     = human_end - human_start,
              viral_pos   = paste(sort(unique(round(hpv_pos))), collapse = ","),
              .groups = "drop") %>%
    mutate(sample_id = sid, related = n_breakpoints >= 2) %>%
    arrange(chr, human_start) %>%
    select(sample_id, cluster_id, hpv_type, chr, human_start, human_end,
           span_bp, n_breakpoints, related, viral_pos)
}

# AA viral intervals for one HN id (empty when AA off).
aa_intervals <- function(sid) {
  if (!isTRUE(params$use_aa) || is.null(aa_raw)) return(tibble(start = double(), end = double()))
  sn <- names(seq2id)[match(sid, seq2id)]
  rows <- aa_raw %>% filter(as.character(`Sample name`) %in% c(sid, sn))
  if (nrow(rows) == 0) return(tibble(start = double(), end = double()))
  locs <- unlist(str_extract_all(rows$Location, "hpv[^']*?:\\d+-\\d+"))
  if (length(locs) == 0) return(tibble(start = double(), end = double()))
  m <- str_match(locs, ":(\\d+)-(\\d+)")
  tibble(start = as.numeric(m[, 2]), end = as.numeric(m[, 3]))
}

# ── Concordance: do searcHPV junctions land on coverage transitions? ──────────
# This does NOT change the state. It reports, per sample:
#   frac_junc_on_cov      fraction of junctions within W bp of a coverage edge
#   n_cov_seg_unexplained sustained coverage segments whose edges no junction marks
#   concordance           a label for quick triage
# HN2408-type cases (junctions sitting away from a broad coverage block) surface
# here as "junctions off coverage" or "partial".
concordance_for <- function(sid, W = params$junc_cov_window_bp) {
  jpos <- junc_pos(sid)
  base <- tibble(n_junc = length(jpos), n_junc_on_cov = NA_integer_,
                 frac_junc_on_cov = NA_real_, n_cov_block = NA_integer_,
                 n_cov_block_unexplained = NA_integer_, concordance = NA_character_)
  if (length(jpos) == 0) return(base)
  zvec <- cov_vec(sid)
  if (length(zvec) == 0) return(mutate(base, concordance = "no coverage"))
  segs  <- cov_blocks(zvec)        # block-level boundaries, not fragmented edges
  edges <- if (nrow(segs)) sort(unique(c(segs$seg_start, segs$seg_end))) else numeric(0)

  junc_on <- vapply(jpos, function(p)
    length(edges) > 0 && min(abs(edges - p)) <= W, logical(1))
  seg_expl <- if (nrow(segs)) vapply(seq_len(nrow(segs)), function(i)
    any(abs(jpos - segs$seg_start[i]) <= W) || any(abs(jpos - segs$seg_end[i]) <= W),
    logical(1)) else logical(0)

  n_on <- sum(junc_on); frac <- n_on / length(jpos)
  n_seg <- nrow(segs); n_unexpl <- if (n_seg) sum(!seg_expl) else 0L
  lab <- if (n_seg == 0) "no coverage segments"
         else if (frac >= 0.5 && n_unexpl == 0) "concordant"
         else if (frac == 0) "junctions off coverage"
         else "partial"
  tibble(n_junc = length(jpos), n_junc_on_cov = n_on, frac_junc_on_cov = frac,
         n_cov_block = n_seg, n_cov_block_unexplained = n_unexpl, concordance = lab)
}

# ── Episomal co-presence (coverage beyond the integrated arc) ─────────────────
# Logic (your call): the integrated viral segment is the span between the
# outermost junctions (the "arc"). HPV coverage OUTSIDE that arc that is not
# dropped out (z >= -z_cut) implies free/episomal HPV in addition to the
# integrated copy. Needs >= 2 junctions to define an arc, and enough genome
# outside the arc to assess. Returns the fraction covered outside and a flag.
#
# CAVEAT: these are cohort-relative z-scores that rarely go deep-negative, so
# outside-arc coverage reads as "present" for most baseline samples. Treat the
# flag as "HPV present beyond the integration", tune episomal_present_frac, and
# read frac_outside_covered as the real quantity.
episomal_for <- function(sid, margin = params$episomal_margin_bp) {
  jp <- junc_pos(sid); z <- cov_vec(sid)
  na <- tibble(frac_outside_covered = NA_real_, n_outside = 0L, episomal_present = NA)
  if (length(jp) < 2 || length(z) == 0) return(na)
  zs  <- roll_mean(z, params$cov_smooth_bp)
  a <- min(jp); b <- max(jp); pos <- seq_along(zs)
  outside <- pos < (a - margin) | pos > (b + margin)
  zo <- zs[outside]; zo <- zo[!is.na(zo)]
  if (length(zo) < 200) return(na)          # arc spans (almost) the whole genome
  frac <- mean(zo >= -params$z_cut)
  tibble(frac_outside_covered = frac, n_outside = length(zo),
         episomal_present = frac >= params$episomal_present_frac)
}
```

------------------------------------------------------------------------

# 5 Classifier

Site count (single vs multiple) is decided by the number of integration
**loci** (`n_loci`), the human-genome clusters from Section 4.1. Two
flanking junctions of one integration sit at one locus and count as a
single site, fixing the earlier over-count where the 200 bp bins split
them into two. Coverage that exceeds the locus count is recorded in
`notes`, not used to change the count. AA-dependent states are only
reachable when `use_aa` is on.

```{r classify-fn}
classify_state <- function(r) {
  hv              <- isTRUE(r$has_hv_junctions)
  cyclic_hpv      <- isTRUE(r$aa_hpv_cyclic)
  cyclic_human    <- isTRUE(r$aa_cyclic_has_human)
  noncontig_flank <- isTRUE(r$flanks_noncontiguous)
  uniform         <- isTRUE(r$cov_uniform)
  excess          <- isTRUE(r$cov_excess)
  # site count comes from integration LOCI (clustered), not raw 200 bp bins, so
  # the two flanking junctions of one integration count as a single site.
  n_loci          <- ifelse(is.na(r$n_loci), 0L, r$n_loci)
  has_cov         <- !is.na(r$cov_frac_gain)

  if (!hv) {
    if (!isTRUE(r$aa_has_hpv) && !has_cov) return("Unclassified")
    if (has_cov && !uniform) return("HPV episomal rearranged")
    return("Episomal only")
  }
  if (cyclic_human || (noncontig_flank && cyclic_hpv))
    return("Human-HPV chimeric episome")
  if (cyclic_hpv || excess) return("Integrated + episomal (mixed)")
  if (n_loci >= 2) return("Integrated clonal, multiple sites")
  if (n_loci == 1) return("Integrated clonal, single site")
  "Unclassified"
}

add_notes <- function(df, aa_enabled = isTRUE(params$use_aa)) {
  df %>%
    mutate(
      cov_extra = coalesce(has_hv_junctions, FALSE) &
                  coalesce(cov_n_event_seg, 0L) > coalesce(n_loci, 0L),
      no_aa = aa_enabled & is.na(aa_has_hpv),
      notes = pmap_chr(
        list(cov_extra, no_aa, cov_n_event_seg, n_loci),
        function(extra, na_aa, nev, ns) {
          msgs <- character()
          if (isTRUE(extra))
            msgs <- c(msgs, sprintf(
              "coverage shows %d event segment(s) vs %d integration locus/loci; extra is coverage-only and NOT counted as a site",
              ifelse(is.na(nev), 0L, nev), ifelse(is.na(ns), 0L, ns)))
          if (isTRUE(na_aa))
            msgs <- c(msgs, "no AmpliconArchitect evidence for this sample")
          if (length(msgs) == 0) "" else paste(msgs, collapse = "; ")
        })
    )
}

classify_table <- function(unified, aa_enabled = isTRUE(params$use_aa)) {
  if (nrow(unified) == 0)
    return(mutate(unified, state = factor(character(), levels = state_levels)))
  states <- pmap_chr(unified, function(...) classify_state(tibble(...)))
  ep <- if ("episomal_present" %in% names(unified)) unified$episomal_present else NA
  unified %>%
    mutate(state_base = states,
           state = if_else(str_starts(state_base, "Integrated clonal") & coalesce(ep, FALSE),
                           paste0(state_base, " + episomal HPV"), state_base),
           state = factor(state, levels = state_levels)) %>%
    add_notes(aa_enabled = aa_enabled)
}
```

------------------------------------------------------------------------

# 6 Classify the OPC cohort

```{r classify-run}
# Related-integration locus clustering FIRST, so the single-vs-multiple decision
# uses loci (one integration locus = one site, even with two flanking junctions),
# not the 200 bp human bins (which split flanking junctions of one event).
conc_ids     <- unified %>% filter(has_hv_junctions) %>% pull(sample_id)
clusters_all <- bind_rows(lapply(conc_ids, integration_clusters))
loci_summary <- if (nrow(clusters_all)) clusters_all %>%
  group_by(sample_id) %>%
  summarise(n_loci = n_distinct(cluster_id),
            n_related_loci = sum(related),
            max_bp_in_locus = max(n_breakpoints), .groups = "drop") else
  tibble(sample_id = character(), n_loci = integer(),
         n_related_loci = integer(), max_bp_in_locus = integer())
unified <- unified %>% left_join(loci_summary, by = "sample_id") %>%
  mutate(n_loci = coalesce(n_loci, 0L), n_related_loci = coalesce(n_related_loci, 0L))

# Episomal co-presence (coverage beyond the integrated arc).
episo_df <- bind_rows(lapply(conc_ids, function(s) mutate(episomal_for(s), sample_id = s)))
if (!nrow(episo_df))
  episo_df <- tibble(sample_id = character(), frac_outside_covered = double(),
                     n_outside = integer(), episomal_present = logical())
unified <- unified %>% left_join(episo_df, by = "sample_id")
if (!"episomal_present" %in% names(unified))
  unified <- mutate(unified, episomal_present = NA, frac_outside_covered = NA_real_, n_outside = 0L)

results <- classify_table(unified)

# Junction-vs-coverage concordance (does NOT change state; flags only).
conc_df <- bind_rows(lapply(conc_ids, function(s) mutate(concordance_for(s), sample_id = s)))
if (nrow(conc_df)) results <- left_join(results, conc_df, by = "sample_id")

state_summary <- results %>% count(state, name = "n", .drop = FALSE) %>% arrange(desc(n))
state_summary %>%
  kable(caption = sprintf("OPC HPV state distribution (use_aa = %s)", params$use_aa)) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)

results %>%
  filter(state != "Unclassified") %>%
  count(state, name = "n", .drop = FALSE) %>%
  ggplot(aes(n, fct_rev(state), fill = state)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = n), hjust = -0.3, size = 4) +
  scale_fill_manual(values = state_palette) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "OPC samples per HPV state", x = "Samples", y = NULL) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.y = element_blank())
```

```{r classify-table-out}
out_tbl <- results %>%
  transmute(
    sample_id, patient, timepoint, state, hpv_type,
    junctions = has_hv_junctions, n_sites, n_loci, n_related_loci,
    episomal_present, frac_outside_covered,
    cov_frac_gain, cov_frac_dip, cov_uniform, cov_excess, cov_n_event_seg,
    concordance, frac_junc_on_cov, n_cov_block, n_cov_block_unexplained,
    aa_hpv_cyclic, aa_cyclic_has_human, notes
  ) %>%
  arrange(patient, timepoint)

write_tsv(out_tbl, file.path(params$out_dir, "hpv_state_OPC.tsv"))

out_tbl %>% head(25) %>%
  kable(caption = "OPC classification (first 25 rows; full table written to results/hpv_state_OPC.tsv)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```

## 6.1 Junction-vs-coverage concordance

This is the recommendation-1 check: it does not change any state, it
only flags whether the searcHPV junctions line up with the coverage
transitions. Use it to gauge how common HN2408-type disagreement is
before deciding whether a deeper change to the integrated-site logic is
worth it.

```{r concordance-summary}
results %>%
  filter(has_hv_junctions) %>%
  count(concordance, name = "n") %>%
  arrange(desc(n)) %>%
  kable(caption = "Concordance among integrated samples (junctions vs coverage edges)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

Integrated samples whose junctions do **not** line up with the coverage
(these are the ones to eyeball, HN2408 should appear here):

```{r discordant-table}
discordant <- results %>%
  filter(has_hv_junctions,
         concordance %in% c("junctions off coverage", "partial")) %>%
  transmute(sample_id, state, n_sites,
            frac_junc_on_cov = round(frac_junc_on_cov, 2),
            n_cov_block, n_cov_block_unexplained) %>%
  arrange(frac_junc_on_cov, desc(n_cov_block_unexplained))

write_tsv(discordant, file.path(params$out_dir, "hpv_discordant_OPC.tsv"))

discordant %>% head(30) %>%
  kable(caption = sprintf("%d discordant integrated samples (full list in results/hpv_discordant_OPC.tsv)",
                          nrow(discordant))) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```

## 6.2 Related-integration loci

Each integrated sample's breakpoints are grouped into human-genome loci:
same HPV type, same chromosome, within `cluster_dist_bp`
(`r format(params$cluster_dist_bp, big.mark = ",")` bp). Breakpoints in
one locus are **related** (the same integration event/region); separate
loci are independent integrations. `n_loci` is how many distinct loci a
sample carries; `related = TRUE` marks a locus holding two or more
breakpoints.

```{r clusters-out}
write_tsv(clusters_all, file.path(params$out_dir, "hpv_integration_clusters_OPC.tsv"))

# How many loci do integrated samples carry?
results %>%
  filter(has_hv_junctions) %>%
  count(n_loci, name = "samples") %>%
  arrange(n_loci) %>%
  kable(caption = "Integration loci per sample (1 = single locus; >1 = multiple independent loci)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

Loci that contain two or more related breakpoints (your "pairs or more
at one locus"), most-clustered first:

```{r related-loci}
clusters_all %>%
  filter(related) %>%
  arrange(desc(n_breakpoints), sample_id) %>%
  transmute(sample_id, locus = cluster_id, hpv_type, chr,
            n_breakpoints, span_bp, viral_pos) %>%
  head(30) %>%
  kable(caption = "Related-integration loci (>= 2 breakpoints within the distance threshold)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```

## 6.3 Episomal co-presence (coverage beyond the integrated arc)

For integrated samples with \>= 2 junctions, `frac_outside_covered` is
the fraction of the HPV genome **outside** the junction-spanned arc that
still shows coverage (z \>= -`r params$z_cut`). At or above
`episomal_present_frac` (`r params$episomal_present_frac`) the sample is
flagged `episomal_present` and its state gets the `+ episomal HPV`
suffix.

Read `frac_outside_covered` as the real signal: because these z-scores
rarely go deep-negative, most baseline samples score high here, so the
binary flag is near-universal. Tune `episomal_present_frac` against this
distribution.

```{r episomal-summary}
results %>%
  filter(has_hv_junctions, !is.na(frac_outside_covered)) %>%
  mutate(bin = cut(frac_outside_covered, breaks = seq(0, 1, 0.1),
                   include.lowest = TRUE)) %>%
  count(bin, name = "samples") %>%
  kable(caption = "Distribution of frac_outside_covered (integrated samples, >=2 junctions)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)

results %>%
  filter(has_hv_junctions) %>%
  count(episomal_present, name = "samples") %>%
  kable(caption = sprintf("episomal_present at threshold %.2f (NA = <2 junctions or arc spans genome)",
                          params$episomal_present_frac)) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

Per-patient view across timepoints (HN as the grouping key):

```{r per-patient}
results %>%
  filter(!is.na(patient)) %>%
  count(patient, state, .drop = FALSE) %>%
  filter(n > 0) %>%
  pivot_wider(names_from = state, values_from = n, values_fill = 0) %>%
  head(20) %>%
  kable(caption = "States per patient across timepoints (first 20 patients)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = TRUE)
```

------------------------------------------------------------------------

# 7 Combined per-sample feature plots

One figure per sample, stacking coverage z-score, searcHPV junctions,
and (when `use_aa`) AA viral intervals on the shared HPV-genome axis.
**Green dotted** verticals are coverage block boundaries; the **junction
points and dashed verticals are coloured by human integration locus**,
so breakpoints sharing a colour are related (same chr, within
`cluster_dist_bp`) and different colours are independent loci. The
**grey band** is the integrated arc (span between the outermost
junctions); coverage outside it drives the episomal call. The caption
lists each locus (`chr:pos`, breakpoint count) and the outside-arc
covered fraction. Discordant samples are drawn first. Each is written as
a PDF named `HPVtracks_<HN id>.pdf` to `params$fig_dir`. Helpers are in
Section 4.1.

```{r plot-fn}
plot_sample_tracks <- function(sid, state_label = "", show_aa = isTRUE(params$use_aa),
                               z_cut = params$z_cut) {
  zvec <- cov_vec(sid)
  jpos <- junc_pos(sid)
  ai   <- aa_intervals(sid)

  trk <- c("Coverage (z-score)", "searcHPV junctions")
  if (show_aa) trk <- c(trk, "AA viral intervals")
  as_trk <- function(x) factor(x, levels = trk)

  cov_df <- if (length(zvec))
    tibble(pos = seq_along(zvec), z = zvec, track = as_trk(trk[1])) else NULL
  segs   <- if (length(zvec)) cov_segments(zvec, cut = z_cut) else tibble()
  seg_df <- if (nrow(segs)) mutate(segs, track = as_trk(trk[1])) else NULL
  thr_df <- tibble(track = as_trk(trk[1]), yint = c(z_cut, -z_cut))
  blk    <- if (length(zvec)) cov_blocks(zvec) else tibble()
  blk_edges <- if (nrow(blk)) sort(unique(c(blk$seg_start, blk$seg_end))) else numeric(0)

  # junctions coloured by their human integration locus (related breakpoints share a colour)
  bp      <- breakpoint_clusters(sid)
  junc_df <- if (nrow(bp)) bp %>% transmute(pos = hpv_pos, locus = cluster_id, y = 1,
                                            track = as_trk(trk[2])) %>% filter(!is.na(pos)) else NULL
  jl_df   <- if (nrow(bp)) bp %>% transmute(pos = hpv_pos, locus = cluster_id) %>%
                                            filter(!is.na(pos)) else NULL
  aa_df   <- if (show_aa && nrow(ai)) mutate(ai, y = 1, track = as_trk("AA viral intervals")) else NULL

  # integrated arc (span between outermost junctions) + episomal metric
  ep <- episomal_for(sid)
  arc_df <- if (length(jpos) >= 2)
    tibble(xmin = min(jpos), xmax = max(jpos), track = as_trk(trk[1])) else NULL

  # caption: list the integration loci (related breakpoints grouped) + episomal
  cl <- integration_clusters(sid)
  loci_cap <- if (nrow(cl)) {
    lab <- sprintf("%s %s (n=%d)", cl$hpv_type, cl$cluster_id, cl$n_breakpoints)
    if (length(lab) > 6) lab <- c(lab[1:6], sprintf("+%d more", length(lab) - 6))
    paste("Integration loci:", paste(lab, collapse = "  |  "))
  } else "Integration loci: none"
  if (!is.na(ep$frac_outside_covered))
    loci_cap <- paste0(loci_cap,
      sprintf("   |   outside-arc covered: %.0f%%%s", 100 * ep$frac_outside_covered,
              if (isTRUE(ep$episomal_present)) " (episomal HPV)" else ""))

  ggplot() +
    { if (!is.null(arc_df))    # integrated arc shaded grey in the coverage panel
        geom_rect(data = arc_df, aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
                  fill = "grey85", alpha = 0.5, inherit.aes = FALSE) } +
    { if (!is.null(seg_df))
        geom_rect(data = seg_df, aes(xmin = seg_start, xmax = seg_end,
                  ymin = -Inf, ymax = Inf, fill = type), alpha = 0.18, inherit.aes = FALSE) } +
    { if (length(blk_edges))   # coverage block boundaries (green dotted)
        geom_vline(xintercept = blk_edges, linetype = "dotted", colour = "#1D9E75", linewidth = 0.5) } +
    { if (!is.null(jl_df))     # searcHPV junctions, coloured by human locus
        geom_vline(data = jl_df, aes(xintercept = pos, colour = locus),
                   linetype = "dashed", linewidth = 0.4) } +
    geom_hline(data = thr_df, aes(yintercept = yint), linetype = "dotted", colour = "grey50") +
    { if (!is.null(cov_df)) geom_line(data = cov_df, aes(pos, z), colour = "#222222", linewidth = 0.35) } +
    { if (!is.null(junc_df)) geom_point(data = junc_df, aes(pos, y, colour = locus), size = 2.8) } +
    { if (!is.null(aa_df)) geom_segment(data = aa_df, aes(x = start, xend = end, y = y, yend = y),
                  colour = "#BA7517", linewidth = 4) } +
    scale_fill_manual(values = c(gain = "#1D9E75", dip = "#D4537E"), name = "Coverage segment") +
    scale_colour_discrete(name = "Integration locus") +
    facet_grid(track ~ ., scales = "free_y", switch = "y") +
    labs(title = paste("Feature tracks —", sid),
         subtitle = if (nzchar(state_label)) paste("State:", state_label) else NULL,
         caption = loci_cap,
         x = "HPV genome position (bp)", y = NULL) +
    theme_minimal(base_size = 12) +
    theme(strip.placement = "outside",
          strip.text.y.left = element_text(angle = 0, face = "bold"),
          panel.grid.minor = element_blank(), legend.position = "bottom")
}
```

```{r choose-and-save}
# integrated samples that also have coverage; discordant ones first so the PDFs
# you review are the suspect calls, then most-sites among the rest.
plot_ids <- results %>%
  filter(has_hv_junctions, !is.na(cov_frac_gain)) %>%
  mutate(is_discordant = concordance %in% c("junctions off coverage", "partial")) %>%
  arrange(desc(is_discordant), frac_junc_on_cov, desc(n_sites)) %>%
  slice_head(n = params$n_plot_samples) %>%
  transmute(sample_id, state = as.character(state))

saved <- character()
for (i in seq_len(nrow(plot_ids))) {
  sid <- plot_ids$sample_id[i]
  p   <- plot_sample_tracks(sid, state_label = plot_ids$state[i])
  fn  <- file.path(params$fig_dir, paste0("HPVtracks_", gsub("[^A-Za-z0-9_]+", "_", sid), ".pdf"))
  ggsave(fn, p, width = 9, height = if (isTRUE(params$use_aa)) 6.5 else 5)
  saved <- c(saved, normalizePath(fn, mustWork = FALSE))
}
cat("Saved per-sample track PDFs:\n"); cat(paste0("  - ", saved), sep = "\n")
```

```{r show-one, fig.width=9, fig.height=5}
if (nrow(plot_ids)) plot_sample_tracks(plot_ids$sample_id[1], plot_ids$state[1])
```

To plot any specific patient/timepoint, call
`plot_sample_tracks("HN2186_Baseline")`.

------------------------------------------------------------------------

# 8 Session info

```{r session-info, echo=FALSE}
sessionInfo()
```
