---
title: "HPV State Classification + Combined Feature Plots"
subtitle: "Integrating searcHPV, AmpliconArchitect, and per-base coverage z-scores"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 3
    code_folding: show
    df_print: paged
params:
  # ── Inputs: exactly the three data types ──────────────────────────────────
  searchhpv_txt: "data/SearcHPV_default_jfz_OPC.trimUMIFastq.txt"   # human-viral junctions
  aa_tsv:        "data/merged_result_tables.tsv"                    # AmpliconArchitect amplicons
  zscore_csv:    "data/20241205_Zscores_stats_nofilter.csv"         # per-base coverage z-scores
  # ── Thresholds (tune here) ────────────────────────────────────────────────
  min_sr_junc:        3      # min srNum to keep a searcHPV breakpoint
  site_bin_bp:        200    # window (bp) to collapse nearby breakpoints into one site
  z_cut:              1.5    # |z| >= this = coverage dip (negative) or gain (positive)
  uniform_frac:       0.05   # if < this fraction of bases deviate, coverage is "uniform"
  excess_gain_frac:   0.50   # if >= this fraction of bases gained, treat as episomal pool
  distant_flank_mb:   10     # human flanks farther apart than this (same chr) = chimeric signal
  focal_cn:           10     # AA max copy number above this = focal amplification
  overlap_window_bp:  200    # window (bp) to call a coverage/junction/AA overlap (Section 7)
  cov_gap_tol:        25     # bp: merge elevated runs separated by gaps <= this into one segment
  cov_min_seg_len:    30     # bp: discard sustained segments shorter than this (noise)
  out_dir:            "results"   # TSV tables go here
  fig_dir:            "."         # per-sample plot PDFs go here (the dir the Rmd runs in)
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6.5)
```

This report classifies each sample into an HPV physical state by lining up
**three** data types, and (new in this version) draws a **combined per-sample
plot** that stacks all three on a shared HPV-genome axis and writes one PDF per
sample to the directory the Rmd is run in.

1. **searcHPV** — human-viral integration junctions (split-read breakpoints).
2. **AmpliconArchitect (AA)** — amplicon architecture from a concatenated
   hg/HPV reference (`Linear`, `Virus`, `Cyclic`, `BFB`, ...), copy number,
   and whether an amplicon contains human + viral sequence together.
3. **Coverage z-scores** — per-base z-score of viral coverage relative to a
   cohort average. Negative = dip, positive = gain.

How coverage is read, per your notes: a **dip** points to integration or (if
effectively zero) deletion; a **gain** points to duplication or integration;
and **multiple separated gain segments** suggest multiple integration
instances. Coverage alone is ambiguous, so it is always combined with the
junction and architecture evidence.

**What changed from v2.** Two things. First, the single-vs-multiple site call
is now driven **only by the number of searcHPV junction sites**. Coverage no
longer silently upgrades a one-junction sample to "multiple sites". Instead,
when coverage shows more events than junctions explain (for example a clear
loss at a second region with no junction and no AA), that is recorded in a
`notes` column but the state stays as the junctions describe it. Second, a
combined three-track figure is produced per sample. See Section 3 for the exact
reasoning and Section 8 for the worked example that motivated the change.

------------------------------------------------------------------------

# 1 Packages

```{r packages}
suppressPackageStartupMessages({
  library(tidyverse)   # dplyr, tidyr, readr, stringr, ggplot2, purrr
  library(knitr)
  library(kableExtra)
})

dir.create(params$out_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(params$fig_dir, showWarnings = FALSE, recursive = TRUE)

# Canonical state labels used throughout
state_levels <- c(
  "Episomal only",
  "HPV episomal rearranged",
  "Human-HPV chimeric episome",
  "Integrated clonal, single site",
  "Integrated clonal, multiple sites",
  "Integrated + episomal (mixed)",
  "Unclassified"
)

state_palette <- c(
  "Episomal only"                     = "#1D9E75",
  "HPV episomal rearranged"           = "#3EC9A7",
  "Human-HPV chimeric episome"        = "#D4537E",
  "Integrated clonal, single site"    = "#534AB7",
  "Integrated clonal, multiple sites" = "#3A2F8F",
  "Integrated + episomal (mixed)"     = "#BA7517",
  "Unclassified"                      = "#888780"
)
```

------------------------------------------------------------------------

# 2 Load and prepare data

Each data type is parsed into a **per-sample feature table** with a shared
`sample_id` key, so the three can be lined up side by side. The parsers below
take an already-read data frame, which lets the same code run on the real
files and on the worked examples in Sections 5 and 8.

## 2.1 searcHPV — junction features

```{r parse-searchpv}
# Real columns: contigName, orientation (hv/vh), genomeInsertionChromosome,
# genomeInsertionPoint, hpvInsertionPoint, SiteConfidence, ContigConfidence,
# srNum, Sample (e.g. "OPC-ReSeq_Cap1_1_S133__HPV33").

parse_searchpv <- function(df, min_sr = params$min_sr_junc,
                           bin_bp = params$site_bin_bp) {
  if (is.null(df) || nrow(df) == 0) {
    return(tibble(sample_id = character(), hpv_type = character(),
                  n_breakpoints = integer(), n_sites = integer(),
                  has_reciprocal = logical(), n_distinct_chr = integer(),
                  chr_coord_range_mb = double(), flanks_noncontiguous = logical(),
                  has_hv_junctions = logical()))
  }

  df %>%
    transmute(
      sample_full = Sample,
      sample_id   = str_remove(Sample, "__HPV\\d+.*$"),
      hpv_type    = str_extract(Sample, "HPV\\d+"),
      orientation = orientation,
      chr_human   = genomeInsertionChromosome,
      pos_human   = suppressWarnings(as.numeric(str_trim(as.character(genomeInsertionPoint)))),
      sr_num      = suppressWarnings(as.numeric(srNum))
    ) %>%
    filter(!is.na(sr_num), sr_num >= min_sr) %>%
    mutate(site_bin = paste(chr_human, floor(pos_human / bin_bp) * bin_bp, sep = ":")) %>%
    group_by(sample_id) %>%
    summarise(
      hpv_type           = first(na.omit(hpv_type)),
      n_breakpoints      = n(),
      n_sites            = n_distinct(site_bin),
      has_reciprocal     = any(orientation == "hv") & any(orientation == "vh"),
      n_distinct_chr     = n_distinct(chr_human),
      chr_coord_range_mb  = if (n_distinct(chr_human) == 1)
                              (max(pos_human, na.rm = TRUE) - min(pos_human, na.rm = TRUE)) / 1e6
                            else NA_real_,
      .groups = "drop"
    ) %>%
    mutate(
      # Non-contiguous human flanks (different chromosomes, or very far apart
      # on the same chromosome) are the chimeric-episome signal.
      flanks_noncontiguous = (n_distinct_chr > 1) |
                             (!is.na(chr_coord_range_mb) &
                                chr_coord_range_mb > params$distant_flank_mb),
      has_hv_junctions     = n_breakpoints > 0
    )
}

shpv_raw <- if (file.exists(params$searchhpv_txt))
  read_tsv(params$searchhpv_txt, show_col_types = FALSE) else NULL
shpv_features <- parse_searchpv(shpv_raw)

cat(sprintf("searcHPV: %d samples with junctions (srNum >= %d).\n",
            nrow(shpv_features), params$min_sr_junc))
head(shpv_features, 8) %>% kable() %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

## 2.2 AmpliconArchitect — architecture features

```{r parse-aa}
# Real columns of interest: "Sample name", "Classification" (Linear/Virus/
# Cyclic/BFB/...), "Location" (e.g. "['chr19:...','hpv16ref_1:0-7909']"),
# "Feature maximum copy number".

parse_aa <- function(df, focal_cn = params$focal_cn) {
  if (is.null(df) || nrow(df) == 0) {
    return(tibble(sample_id = character(), aa_has_hpv = logical(),
                  aa_hpv_cyclic = logical(), aa_cyclic_has_human = logical(),
                  aa_max_cn = double(), aa_focal_amp = logical()))
  }

  df %>%
    transmute(
      sample_id     = `Sample name`,
      classification = Classification,
      location       = Location,
      max_cn         = suppressWarnings(as.numeric(`Feature maximum copy number`)),
      has_hpv_loc    = str_detect(tolower(coalesce(Location, "")), "hpv"),
      has_human_loc  = str_detect(coalesce(Location, ""), "chr"),
      is_cyclic      = str_detect(tolower(coalesce(Classification, "")), "cyclic"),
      is_virus       = str_detect(tolower(coalesce(Classification, "")), "virus")
    ) %>%
    group_by(sample_id) %>%
    summarise(
      aa_has_hpv          = any(has_hpv_loc | is_virus),
      # HPV present inside a cyclic (circular) amplicon = episomal / ecDNA signal
      aa_hpv_cyclic       = any(is_cyclic & has_hpv_loc),
      # cyclic amplicon carrying BOTH human and viral sequence = chimeric ecDNA
      aa_cyclic_has_human = any(is_cyclic & has_hpv_loc & has_human_loc),
      aa_max_cn           = suppressWarnings(max(max_cn, na.rm = TRUE)),
      .groups = "drop"
    ) %>%
    mutate(
      aa_max_cn    = ifelse(is.finite(aa_max_cn), aa_max_cn, NA_real_),
      aa_focal_amp = !is.na(aa_max_cn) & aa_max_cn >= focal_cn
    )
}

aa_raw <- if (file.exists(params$aa_tsv))
  read_tsv(params$aa_tsv, show_col_types = FALSE) else NULL
aa_features <- parse_aa(aa_raw)

cat(sprintf("AmpliconArchitect: %d samples.\n", nrow(aa_features)))
head(aa_features, 8) %>% kable() %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

## 2.3 Coverage z-scores — profile features

```{r parse-cov}
# Real layout: column 1 = "file.name" (sample), remaining columns = per-base
# positions (1,2,3,...) holding the z-score of coverage vs the cohort average.
#
# v3 adds cov_n_dip_seg and cov_n_event_seg (gain + dip). These count separated
# coverage events and are used ONLY for the informational notes in Section 4 —
# they never change the site count, which comes from junctions.

parse_cov <- function(df, z = params$z_cut,
                      uniform_frac = params$uniform_frac,
                      excess_frac  = params$excess_gain_frac) {
  if (is.null(df) || nrow(df) == 0) {
    return(tibble(sample_id = character(), cov_frac_gain = double(),
                  cov_frac_dip = double(), cov_mean_z = double(),
                  cov_uniform = logical(), cov_n_gain_seg = integer(),
                  cov_n_dip_seg = integer(), cov_n_event_seg = integer(),
                  cov_excess = logical()))
  }

  id_col <- names(df)[1]                       # "file.name"
  ids    <- df[[id_col]]
  mat    <- as.matrix(df[, setdiff(names(df), id_col)])
  storage.mode(mat) <- "double"

  frac_gain <- rowMeans(mat >= z,  na.rm = TRUE)
  frac_dip  <- rowMeans(mat <= -z, na.rm = TRUE)
  mean_z    <- rowMeans(mat,        na.rm = TRUE)

  # count separated runs of contiguous gain / dip = candidate event instances
  count_seg <- function(v, up) {
    flag <- if (up) (!is.na(v) & v >= z) else (!is.na(v) & v <= -z)
    if (!any(flag)) return(0L)
    sum(rle(flag)$values)
  }
  n_gain_seg <- apply(mat, 1, count_seg, up = TRUE)
  n_dip_seg  <- apply(mat, 1, count_seg, up = FALSE)

  tibble(
    sample_id       = ids,
    cov_frac_gain   = frac_gain,
    cov_frac_dip    = frac_dip,
    cov_mean_z      = mean_z,
    cov_uniform     = (frac_gain + frac_dip) < uniform_frac,
    cov_n_gain_seg  = as.integer(n_gain_seg),
    cov_n_dip_seg   = as.integer(n_dip_seg),
    cov_n_event_seg = as.integer(n_gain_seg + n_dip_seg),
    cov_excess      = frac_gain >= excess_frac
  )
}

cov_raw <- if (file.exists(params$zscore_csv))
  read_csv(params$zscore_csv, show_col_types = FALSE) else NULL
cov_features <- parse_cov(cov_raw)

cat(sprintf("Coverage: %d samples profiled.\n", nrow(cov_features)))
head(cov_features, 8) %>% kable(digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

## 2.4 Unified feature table

The three feature tables are joined on `sample_id`. The real input files come
from three different cohorts (their sample naming does not overlap), so on the
real data each row typically carries only one data type. That is expected; the
join is the mechanism, and Sections 5 and 8 supply aligned examples where all
three line up per sample.

```{r unify}
build_unified <- function(shpv, aa, cov) {
  ids <- unique(c(shpv$sample_id, aa$sample_id, cov$sample_id))
  tibble(sample_id = ids) %>%
    left_join(shpv, by = "sample_id") %>%
    left_join(aa,   by = "sample_id") %>%
    left_join(cov,  by = "sample_id") %>%
    mutate(
      has_hv_junctions = coalesce(has_hv_junctions, FALSE),
      n_breakpoints    = coalesce(n_breakpoints, 0L),
      n_sites          = coalesce(n_sites, 0L)
    )
}

unified_real <- build_unified(shpv_features, aa_features, cov_features)
cat(sprintf("Unified real table: %d samples (mostly single-data-type rows).\n",
            nrow(unified_real)))
```

------------------------------------------------------------------------

# 3 Decision logic

Each sample is resolved by combining the three evidence streams. The order
follows the classification diagram (junctions first, then architecture, then
coverage as a modifier).

| Step | Question | Evidence used |
|------|----------|---------------|
| Q1 | Human-viral junctions present? | searcHPV `has_hv_junctions` |
| Q2 | (no junctions) Coverage uniform? | z-score profile `cov_uniform` |
| Q3 | (junctions) Chimeric episome? | AA cyclic-with-human OR non-contiguous human flanks |
| Q4 | Single vs multiple integration sites? | searcHPV `n_sites` **(authoritative)** |
| Q5 | Episomal modifier? | AA HPV-cyclic OR broad excess coverage gain |

Coverage reading inside this logic: with **no junctions**, a uniform profile is
episomal; internal dips/peaks within the viral genome (non-uniform, no human
sequence) indicate **rearrangement**. With **junctions present**, broad excess
gain beyond what the junctions explain, or a circular HPV amplicon, flags a
coexisting **episomal pool**.

**Site counting (Q4) is decided by junctions, not coverage.** This is the v2
behaviour fix. In v2, a sample with one junction site was upgraded to "multiple
sites" whenever coverage showed two or more gain blocks, and coverage *losses*
were not counted at all. That meant the case you raised — a clear coverage loss
plus an integration junction at one site, with a second region of coverage
change that has no junction and no AA — behaved inconsistently. v3 makes the
rule explicit: **the searcHPV junction site count sets single vs multiple.**
Coverage events that exceed the junction count (gains or losses) are reported in
the `notes` column as coverage-only signal, because on their own they are
ambiguous (a loss can be integration *or* deletion, and there is no junction or
AA to disambiguate). The worked example is in Section 8.

> Subclonal integration is intentionally **not** called here. Its definition
> depends on junction VAF versus tumour purity, and purity is outside these
> three data types, so it is deferred.

```{r classify-fn}
classify_state <- function(r) {

  hv              <- isTRUE(r$has_hv_junctions)
  cyclic_hpv      <- isTRUE(r$aa_hpv_cyclic)
  cyclic_human    <- isTRUE(r$aa_cyclic_has_human)
  noncontig_flank <- isTRUE(r$flanks_noncontiguous)
  uniform         <- isTRUE(r$cov_uniform)
  excess          <- isTRUE(r$cov_excess)
  n_sites         <- ifelse(is.na(r$n_sites), 0L, r$n_sites)
  has_cov         <- !is.na(r$cov_frac_gain)

  # ── Q1: no human-viral junctions → episomal family ──────────────────────
  if (!hv) {
    # no junctions, no AA HPV, and no coverage row = nothing to go on
    if (!isTRUE(r$aa_has_hpv) && !has_cov) return("Unclassified")
    # Q2: uniform coverage = intact episome; non-uniform within HPV = rearranged
    if (has_cov && !uniform) return("HPV episomal rearranged")
    return("Episomal only")
  }

  # ── Junctions present ───────────────────────────────────────────────────
  # Q3: chimeric episome — viral+human on a CIRCULAR molecule. A chimeric
  #     episome is ecDNA (AA cyclic) carrying human + viral sequence. Multi-site
  #     integration also produces junctions on multiple chromosomes, so
  #     non-contiguous flanks only count as chimeric when the molecule is also
  #     circular (otherwise they are just separate linear integration sites).
  if (cyclic_human || (noncontig_flank && cyclic_hpv))
    return("Human-HPV chimeric episome")

  # Q5: episomal modifier — circular HPV amplicon, or coverage far in excess of
  #     what the junctions alone would produce (a free episomal pool).
  if (cyclic_hpv || excess) return("Integrated + episomal (mixed)")

  # Q4: clonal integration — single vs multiple sites, decided by junctions.
  #     Coverage is NOT used to override the site count (see notes column).
  if (n_sites >= 2) return("Integrated clonal, multiple sites")
  if (n_sites == 1) return("Integrated clonal, single site")
  "Unclassified"
}
```

------------------------------------------------------------------------

# 4 Apply the classifier

```{r apply-fn}
# Informational notes. These never change the state; they surface where the
# evidence streams disagree so a reviewer can see it.
add_notes <- function(df) {
  df %>%
    mutate(
      no_aa     = is.na(aa_has_hpv),
      cov_extra = coalesce(has_hv_junctions, FALSE) &
                  coalesce(cov_n_event_seg, 0L) > coalesce(n_sites, 0L),
      notes = pmap_chr(
        list(no_aa, cov_extra, cov_n_event_seg, n_sites),
        function(na_aa, extra, nev, ns) {
          msgs <- character()
          if (isTRUE(extra))
            msgs <- c(msgs, sprintf(
              "coverage shows %d event segment(s) vs %d junction site(s); extra event is coverage-only (no junction/AA) and is NOT counted as a site",
              nev, ns))
          if (isTRUE(na_aa))
            msgs <- c(msgs, "no AmpliconArchitect evidence")
          if (length(msgs) == 0) "" else paste(msgs, collapse = "; ")
        })
    )
}

classify_table <- function(unified) {
  if (nrow(unified) == 0)
    return(mutate(unified, state = factor(character(), levels = state_levels)))
  states <- pmap_chr(unified, function(...) classify_state(tibble(...)))
  unified %>%
    mutate(state = factor(states, levels = state_levels)) %>%
    add_notes()
}
```

------------------------------------------------------------------------

# 5 Synthetic aligned test set

Six engineered samples where all three data types line up per sample, each built
to land in a known state. The same parsers and classifier from above are reused,
so this also exercises the loaders end to end.

```{r synth-data}
# ── Synthetic searcHPV rows ────────────────────────────────────────────────
synth_shpv <- tribble(
  ~Sample,                ~orientation, ~genomeInsertionChromosome, ~genomeInsertionPoint, ~hpvInsertionPoint, ~SiteConfidence,        ~ContigConfidence,        ~srNum,
  "S_single__HPV16",      "hv",         "chr13",                    34110000,              1200,               "High Confidence Site", "High Confidence Contig", 120,
  "S_single__HPV16",      "vh",         "chr13",                    34110150,              6500,               "High Confidence Site", "High Confidence Contig", 110,
  "S_multi__HPV16",       "hv",         "chr8",                     128700000,             900,                "High Confidence Site", "High Confidence Contig", 90,
  "S_multi__HPV16",       "vh",         "chr8",                     128700120,             5200,               "High Confidence Site", "High Confidence Contig", 85,
  "S_multi__HPV16",       "hv",         "chr3",                     61500000,              3100,               "High Confidence Site", "High Confidence Contig", 80,
  "S_multi__HPV16",       "vh",         "chr3",                     61500140,              7100,               "High Confidence Site", "High Confidence Contig", 78,
  "S_chim__HPV16",        "hv",         "chr7",                     55000000,              1000,               "High Confidence Site", "High Confidence Contig", 60,
  "S_chim__HPV16",        "vh",         "chr12",                    25000000,              6000,               "High Confidence Site", "High Confidence Contig", 58,
  "S_mixed__HPV16",       "hv",         "chr5",                     1295000,               1500,               "High Confidence Site", "High Confidence Contig", 70,
  "S_mixed__HPV16",       "vh",         "chr5",                     1295160,               6800,               "High Confidence Site", "High Confidence Contig", 66
  # S_epi and S_rearr deliberately have NO human-viral junctions
)

# ── Synthetic AA rows ──────────────────────────────────────────────────────
synth_aa <- tribble(
  ~`Sample name`, ~Classification, ~Location,                                   ~`Feature maximum copy number`,
  "S_epi",        "Virus",         "['hpv16ref_1:0-7909']",                     8,
  "S_rearr",      "Virus",         "['hpv16ref_1:0-7909']",                     12,
  "S_chim",       "Cyclic",        "['chr7:55000000-55050000','hpv16ref_1:0-7909','chr12:25000000-25040000']", 45,
  "S_single",     "Linear",        "['chr13:34100000-34120000','hpv16ref_1:0-7909']", 6,
  "S_multi",      "Linear",        "['chr8:128690000-128710000','hpv16ref_1:0-7909']", 5,
  "S_multi",      "Linear",        "['chr3:61490000-61510000']",                4,
  "S_mixed",      "Linear",        "['chr5:1290000-1300000','hpv16ref_1:0-7909']", 7,
  "S_mixed",      "Cyclic",        "['hpv16ref_1:0-7909']",                     30
)

# ── Synthetic coverage z-score matrix (200 bases) ──────────────────────────
set.seed(1)
n_pos <- 200
mk_profile <- function(kind) {
  z <- rnorm(n_pos, 0, 0.3)                       # baseline noise
  if (kind == "uniform")      z <- z
  if (kind == "internal_dip") z[60:100] <- z[60:100] - 4      # dip inside HPV
  if (kind == "broad_gain")   z <- z + 3                       # whole-genome amplified
  if (kind == "one_gain")     z[80:110] <- z[80:110] + 4       # one gained block
  if (kind == "two_gains")  { z[30:55]  <- z[30:55]  + 4
                              z[130:160] <- z[130:160] + 4 }    # two gained blocks
  if (kind == "excess")     { z <- z + 2; z[90:120] <- z[90:120] + 3 }  # baseline lift + bump
  z
}
synth_cov <- bind_rows(
  c(list(file.name = "S_epi"),    setNames(as.list(mk_profile("uniform")),      1:n_pos)),
  c(list(file.name = "S_rearr"),  setNames(as.list(mk_profile("internal_dip")), 1:n_pos)),
  c(list(file.name = "S_chim"),   setNames(as.list(mk_profile("broad_gain")),   1:n_pos)),
  c(list(file.name = "S_single"), setNames(as.list(mk_profile("one_gain")),     1:n_pos)),
  c(list(file.name = "S_multi"),  setNames(as.list(mk_profile("two_gains")),    1:n_pos)),
  c(list(file.name = "S_mixed"),  setNames(as.list(mk_profile("excess")),       1:n_pos))
)
```

```{r synth-classify}
unified_synth <- build_unified(
  parse_searchpv(synth_shpv),
  parse_aa(synth_aa),
  parse_cov(synth_cov)
)

results_synth <- classify_table(unified_synth)
```

------------------------------------------------------------------------

# 6 Final classification

One row per sample with its assigned HPV state, the evidence that drove it, and
any disagreement notes.

```{r final-table}
final_df <- results_synth %>%
  transmute(
    sample_id,
    state,
    hpv_type,
    junctions      = has_hv_junctions,
    n_sites,
    cov_n_event_seg,
    aa_hpv_cyclic,
    aa_cyclic_has_human,
    flanks_noncontiguous,
    cov_uniform,
    cov_excess,
    notes
  ) %>%
  arrange(state)

write_tsv(final_df, file.path(params$out_dir, "hpv_state_classification.tsv"))

final_df %>%
  kable(caption = "HPV state classification (synthetic aligned test set)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE)
```

```{r final-summary}
final_df %>%
  count(state, name = "n", .drop = FALSE) %>%
  ggplot(aes(n, fct_rev(state), fill = state)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = n), hjust = -0.4, size = 4) +
  scale_fill_manual(values = state_palette) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Samples per HPV state", x = "Samples", y = NULL) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.y = element_blank())
```

The same `classify_table()` runs on the real unified table; rows there carry
only partial evidence (different cohorts), so most resolve to `Episomal only`
or `Unclassified` until the three assays are run on shared samples.

```{r real-classify}
results_real <- classify_table(unified_real)
results_real %>% count(state, name = "n", .drop = FALSE) %>%
  kable(caption = "Real data (partial evidence per cohort)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

------------------------------------------------------------------------

# 7 Concordance helpers (shared by plots and tables)

These functions place the three data types on the shared HPV axis. The plot in
Section 8 and the concordance table both use them.

A coverage segment edge is the boundary of a **sustained** run: bases
continuously beyond |z| >= `r params$z_cut` on average. Brief sub-threshold
flickers inside a run are bridged (`cov_gap_tol`) and short noise spikes are
dropped (`cov_min_seg_len`), so a single integrated viral block yields exactly
two edges (start and end), not a scatter of crossings.

```{r concord-fns}
seg_from_flag <- function(flag, gap_tol) {
  idx <- which(flag)
  if (length(idx) == 0) return(list())
  brk <- c(0L, which(diff(idx) > (gap_tol + 1L)), length(idx))
  lapply(seq_len(length(brk) - 1L), function(j) {
    grp <- idx[(brk[j] + 1L):brk[j + 1L]]
    c(min(grp), max(grp))
  })
}

cov_segments <- function(zvec, cut = params$z_cut,
                         gap_tol = params$cov_gap_tol,
                         min_len = params$cov_min_seg_len) {
  out <- list()
  for (dir in c(1L, -1L)) {
    flag <- if (dir == 1L) zvec >= cut else zvec <= -cut
    for (m in seg_from_flag(flag, gap_tol)) {
      s <- m[1]; e <- m[2]
      if (e - s + 1L < min_len) next
      mz <- mean(zvec[s:e])
      if ((dir == 1L && mz >= cut) || (dir == -1L && mz <= -cut)) {
        out[[length(out) + 1L]] <- tibble(seg_start = s, seg_end = e,
                                          type = if (dir == 1L) "gain" else "dip",
                                          mean_z = mz, len = e - s + 1L)
      }
    }
  }
  if (length(out) == 0)
    return(tibble(seg_start = integer(), seg_end = integer(),
                  type = character(), mean_z = double(), len = integer()))
  bind_rows(out) %>% arrange(seg_start)
}

cov_edges <- function(zvec, ...) {
  segs <- cov_segments(zvec, ...)
  if (nrow(segs) == 0) return(integer())
  sort(unique(c(segs$seg_start, segs$seg_end)))
}

# searcHPV viral coordinate of each junction for one sample.
junction_viral_pos <- function(df, sid) {
  df %>%
    mutate(sample_id = str_remove(Sample, "__HPV\\d+.*$"),
           hpv_pos   = suppressWarnings(as.numeric(str_trim(as.character(hpvInsertionPoint))))) %>%
    filter(sample_id == sid, !is.na(hpv_pos)) %>% pull(hpv_pos)
}

# AA viral intervals (start/end pairs) for one sample, for plotting bars.
aa_viral_intervals <- function(df, sid) {
  rows <- df %>% filter(`Sample name` == sid)
  if (nrow(rows) == 0) return(tibble(start = double(), end = double()))
  locs <- unlist(str_extract_all(rows$Location, "hpv[^']*?:\\d+-\\d+"))
  if (length(locs) == 0) return(tibble(start = double(), end = double()))
  m <- str_match(locs, ":(\\d+)-(\\d+)")
  tibble(start = as.numeric(m[, 2]), end = as.numeric(m[, 3]))
}

# AA viral boundary positions (flat) for concordance anchoring.
aa_viral_boundaries <- function(df, sid) {
  ai <- aa_viral_intervals(df, sid)
  if (nrow(ai) == 0) return(numeric(0))
  sort(unique(c(ai$start, ai$end)))
}

# Cluster all anchors within W bp and tag which data types support each region.
build_concordance <- function(sid, zvec, shpv_df, aa_df,
                              W = params$overlap_window_bp) {
  anchors <- bind_rows(
    tibble(source = "coverage", pos = cov_edges(zvec)),
    tibble(source = "junction", pos = junction_viral_pos(shpv_df, sid)),
    tibble(source = "AA",       pos = aa_viral_boundaries(aa_df, sid))
  ) %>% filter(!is.na(pos)) %>% arrange(pos)
  if (nrow(anchors) == 0) return(tibble())

  grp <- integer(nrow(anchors)); g <- 0L; last <- -Inf
  for (i in seq_len(nrow(anchors))) {
    if (anchors$pos[i] - last > W) g <- g + 1L
    grp[i] <- g; last <- anchors$pos[i]
  }
  anchors$region <- grp

  anchors %>%
    group_by(region) %>%
    summarise(
      sample_id = sid,
      hpv_start = min(pos), hpv_end = max(pos),
      coverage  = any(source == "coverage"),
      junction  = any(source == "junction"),
      AA        = any(source == "AA"),
      .groups   = "drop"
    ) %>%
    mutate(n_types = coverage + junction + AA,
           support = case_when(n_types == 3 ~ "all three",
                               n_types == 2 ~ "two of three",
                               TRUE         ~ "single")) %>%
    select(sample_id, region, hpv_start, hpv_end,
           coverage, junction, AA, n_types, support)
}
```

------------------------------------------------------------------------

# 8 Combined per-sample feature plot

This is the new deliverable: one figure per sample that stacks all three data
types on the shared HPV-genome axis.

- **Top track** — coverage z-score line, with dotted threshold lines at
  ±`r params$z_cut`. Sustained gain segments are shaded green, sustained loss
  segments pink.
- **Middle track** — searcHPV junction viral positions (points). The dashed
  vertical guides extend through all three panels so you can eyeball alignment.
- **Bottom track** — AmpliconArchitect viral intervals (orange bars).
- A light green vertical band marks regions where **all three** agree (within
  `r params$overlap_window_bp` bp).

Each plot is written as a **PDF** to `params$fig_dir`
(default `"."`, i.e. the directory the Rmd runs in), named
`HPVtracks_<sample>.pdf`.

## 8.1 The classification question, worked

You asked: *if there is a clear loss of coverage and an integration detected at
one site but not another, with no AA information, what is it classified as?*

Sample `S_cov2nd` below is exactly that case: a single searcHPV junction site on
chr13 (with a coverage gain at the integrated viral block), **plus a clear
coverage loss at a second region** that has **no junction and no AA row**.

The answer: it is classified as **"Integrated clonal, single site."** The
junction site count is authoritative, so the lone junction site sets the state.
The second-region coverage loss is real and is shown on the plot and recorded in
the `notes` column ("coverage shows N event segments vs 1 junction site; extra
event is coverage-only..."), but because a coverage loss with no junction and no
AA is ambiguous (it could be a second integration *or* a deletion), it is not
promoted to a second site. The note makes the disagreement visible for review.

```{r demo-data}
hpv_len <- 7906
mk_full <- function(blocks, len = hpv_len, sd = 0.3, seed = 42) {
  set.seed(seed)
  z <- rnorm(len, 0, sd)
  for (b in blocks) z[b[1]:b[2]] <- z[b[1]:b[2]] + b[3]
  z
}

demo_z <- list(
  S_single = mk_full(list(c(1200, 6500, 4)),               seed = 11),
  S_multi  = mk_full(list(c(900, 5200, 4), c(3100, 7100, 4)), seed = 12),
  # the worked case: one junction site (gain at integration) + a clear LOSS
  # at a second region with no junction and no AA.
  S_cov2nd = mk_full(list(c(1200, 2500, 4), c(5000, 6000, -4)), seed = 13)
)

demo_shpv <- tribble(
  ~Sample,            ~orientation, ~genomeInsertionChromosome, ~genomeInsertionPoint, ~hpvInsertionPoint, ~SiteConfidence,        ~ContigConfidence,        ~srNum,
  "S_single__HPV16",  "hv",         "chr13",                    34110000,              1200,               "High Confidence Site", "High Confidence Contig", 120,
  "S_single__HPV16",  "vh",         "chr13",                    34110150,              6500,               "High Confidence Site", "High Confidence Contig", 110,
  "S_multi__HPV16",   "hv",         "chr8",                     128700000,             900,                "High Confidence Site", "High Confidence Contig", 90,
  "S_multi__HPV16",   "vh",         "chr8",                     128700120,             5200,               "High Confidence Site", "High Confidence Contig", 85,
  "S_multi__HPV16",   "hv",         "chr3",                     61500000,              3100,               "High Confidence Site", "High Confidence Contig", 80,
  "S_multi__HPV16",   "vh",         "chr3",                     61500140,              7100,               "High Confidence Site", "High Confidence Contig", 78,
  "S_cov2nd__HPV16",  "hv",         "chr13",                    34110000,              1300,               "High Confidence Site", "High Confidence Contig", 70,
  "S_cov2nd__HPV16",  "vh",         "chr13",                    34110150,              2400,               "High Confidence Site", "High Confidence Contig", 66
)

aa_demo <- tribble(
  ~`Sample name`, ~Classification, ~Location,                                            ~`Feature maximum copy number`,
  "S_single",     "Linear",        "['chr13:34100000-34120000','hpv16ref_1:1200-6500']", 6,
  "S_multi",      "Linear",        "['chr8:128690000-128710000','hpv16ref_1:900-5200']", 5,
  "S_multi",      "Linear",        "['chr3:61490000-61510000','hpv16ref_1:3100-7100']",  4
  # S_cov2nd intentionally absent: no AA evidence
)
```

```{r demo-classify}
demo_cov_wide <- imap_dfr(demo_z, function(z, nm) {
  bind_cols(tibble(file.name = nm),
            as_tibble_row(setNames(as.list(as.numeric(z)), as.character(seq_along(z)))))
})

demo_unified <- build_unified(
  parse_searchpv(demo_shpv),
  parse_aa(aa_demo),
  parse_cov(demo_cov_wide)
)
demo_results <- classify_table(demo_unified)

demo_results %>%
  transmute(sample_id, state, n_sites,
            cov_n_gain_seg, cov_n_dip_seg, cov_n_event_seg, no_aa, notes) %>%
  arrange(sample_id) %>%
  kable(caption = "Worked examples, including the coverage-loss-at-second-site case (S_cov2nd)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```

## 8.2 Plot function

```{r plot-fn}
plot_sample_tracks <- function(sid, zvec, shpv_df, aa_df,
                               z_cut   = params$z_cut,
                               gap_tol = params$cov_gap_tol,
                               min_len = params$cov_min_seg_len,
                               W       = params$overlap_window_bp) {

  trk <- c("Coverage (z-score)", "searcHPV junctions", "AA viral intervals")
  as_trk <- function(x) factor(x, levels = trk)

  # ── coverage track ──
  cov_df <- tibble(pos = seq_along(zvec), z = as.numeric(zvec), track = as_trk(trk[1]))
  segs   <- cov_segments(zvec, cut = z_cut, gap_tol = gap_tol, min_len = min_len)
  seg_df <- if (nrow(segs)) mutate(segs, track = as_trk(trk[1])) else NULL
  thr_df <- tibble(track = as_trk(trk[1]), yint = c(z_cut, -z_cut))

  # ── junction track ──
  jpos    <- junction_viral_pos(shpv_df, sid)
  junc_df <- if (length(jpos)) tibble(pos = jpos, y = 1, track = as_trk(trk[2])) else NULL

  # ── AA track ──
  ai     <- aa_viral_intervals(aa_df, sid)
  aa_df2 <- if (nrow(ai)) mutate(ai, y = 1, track = as_trk(trk[3])) else NULL

  # ── all-three concordance band (drawn in every panel) ──
  conc <- tryCatch(build_concordance(sid, zvec, shpv_df, aa_df, W = W),
                   error = function(e) tibble())
  all3 <- if (nrow(conc)) filter(conc, n_types == 3) else conc

  ggplot() +
    { if (nrow(all3))
        geom_rect(data = all3,
                  aes(xmin = hpv_start - W/2, xmax = hpv_end + W/2,
                      ymin = -Inf, ymax = Inf),
                  fill = "#e6f4ea", inherit.aes = FALSE) } +
    { if (!is.null(seg_df))
        geom_rect(data = seg_df,
                  aes(xmin = seg_start, xmax = seg_end, ymin = -Inf, ymax = Inf,
                      fill = type),
                  alpha = 0.18, inherit.aes = FALSE) } +
    { if (length(jpos))
        geom_vline(xintercept = jpos, linetype = "dashed",
                   colour = "#534AB7", linewidth = 0.4) } +
    geom_hline(data = thr_df, aes(yintercept = yint),
               linetype = "dotted", colour = "grey50") +
    geom_line(data = cov_df, aes(pos, z), colour = "#222222", linewidth = 0.4) +
    { if (!is.null(junc_df))
        geom_point(data = junc_df, aes(pos, y), colour = "#534AB7", size = 2.8) } +
    { if (!is.null(aa_df2))
        geom_segment(data = aa_df2, aes(x = start, xend = end, y = y, yend = y),
                     colour = "#BA7517", linewidth = 4) } +
    scale_fill_manual(values = c(gain = "#1D9E75", dip = "#D4537E"),
                      name = "Coverage segment") +
    facet_grid(track ~ ., scales = "free_y", switch = "y") +
    labs(title = paste("Integrated feature tracks —", sid),
         subtitle = sprintf("HPV axis (bp). All-three-agree regions shaded green. State: %s",
                            as.character(classify_state(
                              demo_results %>% filter(sample_id == sid) %>% slice(1)))),
         x = "HPV genome position (bp)", y = NULL) +
    theme_minimal(base_size = 12) +
    theme(strip.placement = "outside",
          strip.text.y.left = element_text(angle = 0, face = "bold"),
          panel.grid.minor = element_blank(),
          legend.position = "bottom")
}
```

## 8.3 Render inline and write one PDF per sample

```{r show-one, fig.width=9, fig.height=6.5}
# Show the worked case inline in the HTML report
plot_sample_tracks("S_cov2nd", demo_z$S_cov2nd, demo_shpv, aa_demo)
```

```{r save-plots}
demo_samples <- names(demo_z)
saved <- character()
for (sid in demo_samples) {
  p  <- plot_sample_tracks(sid, demo_z[[sid]], demo_shpv, aa_demo)
  fn <- file.path(params$fig_dir, paste0("HPVtracks_", sid, ".pdf"))
  ggsave(fn, p, width = 9, height = 6.5)   # PDF device chosen from extension
  saved <- c(saved, normalizePath(fn, mustWork = FALSE))
}
cat("Saved per-sample combined-feature PDFs:\n")
cat(paste0("  - ", saved), sep = "\n")
```

To plot your real samples, call `plot_sample_tracks()` with that sample's
z-score vector, the searcHPV data frame, and the AA data frame, then `ggsave()`
the same way. The function works on any single sample id present in those
inputs.

------------------------------------------------------------------------

# 9 Coordinate concordance table (where the three agree)

The per-sample plot is the visual view; this is the tabular view, listing every
region on the HPV axis and how many data types support it.

```{r concord-table}
concordance_tbl <- bind_rows(lapply(demo_samples, function(sid)
  build_concordance(sid, demo_z[[sid]], demo_shpv, aa_demo)))

write_tsv(concordance_tbl, file.path(params$out_dir, "hpv_axis_concordance.tsv"))

concordance_tbl %>%
  kable(caption = "HPV-axis overlap regions (coverage edge / junction / AA boundary)",
        col.names = c("Sample", "Region", "HPV start", "HPV end",
                      "Coverage", "Junction", "AA", "N types", "Support")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) %>%
  row_spec(which(concordance_tbl$n_types == 3), background = "#e6f4ea")
```

A region supported by **all three** is the strongest integration call from
these inputs. For `S_cov2nd`, note that the second-region coverage loss appears
here as a **coverage-only** region (n_types = 1), consistent with it not being
counted as a junction site.

------------------------------------------------------------------------

# 10 Session info

```{r session-info, echo=FALSE}
sessionInfo()
```
