---
title: "HPV State Classification + Coordinate Concordance"
subtitle: "Integrating searcHPV, AmpliconArchitect, and per-base coverage z-scores"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 3
    code_folding: show
    df_print: paged
params:
  # ── Inputs: exactly the three data types ──────────────────────────────────
  searchhpv_txt: "data/SearcHPV_default_jfz_OPC.trimUMIFastq.txt"   # human-viral junctions
  aa_tsv:        "data/merged_result_tables.tsv"                    # AmpliconArchitect amplicons
  zscore_csv:    "data/20241205_Zscores_stats_nofilter.csv"         # per-base coverage z-scores
  # ── Thresholds (tune here) ────────────────────────────────────────────────
  min_sr_junc:        3      # min srNum to keep a searcHPV breakpoint
  site_bin_bp:        200    # window (bp) to collapse nearby breakpoints into one site
  z_cut:              1.5    # |z| >= this = coverage dip (negative) or gain (positive)
  uniform_frac:       0.05   # if < this fraction of bases deviate, coverage is "uniform"
  excess_gain_frac:   0.50   # if >= this fraction of bases gained, treat as episomal pool
  distant_flank_mb:   10     # human flanks farther apart than this (same chr) = chimeric signal
  focal_cn:           10     # AA max copy number above this = focal amplification
  overlap_window_bp:  200    # window (bp) to call a coverage/junction/AA overlap (Section 7)
  cov_gap_tol:        25     # bp: merge elevated runs separated by gaps <= this into one segment
  cov_min_seg_len:    30     # bp: discard sustained segments shorter than this (noise)
  out_dir:            "results"
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5)
```

This report classifies each sample into an HPV physical state by lining up
**three** data types:

1. **searcHPV** — human-viral integration junctions (split-read breakpoints).
2. **AmpliconArchitect (AA)** — amplicon architecture from a concatenated
   hg/HPV reference (`Linear`, `Virus`, `Cyclic`, `BFB`, ...), copy number,
   and whether an amplicon contains human + viral sequence together.
3. **Coverage z-scores** — per-base z-score of viral coverage relative to a
   cohort average. Negative = dip, positive = gain.

How coverage is read, per your notes: a **dip** points to integration or (if
effectively zero) deletion; a **gain** points to duplication or integration;
and **multiple separated gain segments** suggest multiple integration
instances. Coverage alone is ambiguous, so it is always combined with the
junction and architecture evidence.

------------------------------------------------------------------------

# 1 Packages

```{r packages}
suppressPackageStartupMessages({
  library(tidyverse)   # dplyr, tidyr, readr, stringr, ggplot2, purrr
  library(knitr)
  library(kableExtra)
})

dir.create(params$out_dir, showWarnings = FALSE, recursive = TRUE)

# Canonical state labels used throughout
state_levels <- c(
  "Episomal only",
  "HPV episomal rearranged",
  "Human-HPV chimeric episome",
  "Integrated clonal, single site",
  "Integrated clonal, multiple sites",
  "Integrated + episomal (mixed)",
  "Unclassified"
)

state_palette <- c(
  "Episomal only"                     = "#1D9E75",
  "HPV episomal rearranged"           = "#3EC9A7",
  "Human-HPV chimeric episome"        = "#D4537E",
  "Integrated clonal, single site"    = "#534AB7",
  "Integrated clonal, multiple sites" = "#3A2F8F",
  "Integrated + episomal (mixed)"     = "#BA7517",
  "Unclassified"                      = "#888780"
)
```

------------------------------------------------------------------------

# 2 Load and prepare data

Each data type is parsed into a **per-sample feature table** with a shared
`sample_id` key, so the three can be lined up side by side. The parsers below
take an already-read data frame, which lets the same code run on the real
files and on the synthetic test set in Section 5.

## 2.1 searcHPV — junction features

```{r parse-searchpv}
# Real columns: contigName, orientation (hv/vh), genomeInsertionChromosome,
# genomeInsertionPoint, hpvInsertionPoint, SiteConfidence, ContigConfidence,
# srNum, Sample (e.g. "OPC-ReSeq_Cap1_1_S133__HPV33").

parse_searchpv <- function(df, min_sr = params$min_sr_junc,
                           bin_bp = params$site_bin_bp) {
  if (is.null(df) || nrow(df) == 0) {
    return(tibble(sample_id = character(), hpv_type = character(),
                  n_breakpoints = integer(), n_sites = integer(),
                  has_reciprocal = logical(), n_distinct_chr = integer(),
                  chr_coord_range_mb = double(), flanks_noncontiguous = logical(),
                  has_hv_junctions = logical()))
  }

  df %>%
    transmute(
      sample_full = Sample,
      sample_id   = str_remove(Sample, "__HPV\\d+.*$"),
      hpv_type    = str_extract(Sample, "HPV\\d+"),
      orientation = orientation,
      chr_human   = genomeInsertionChromosome,
      pos_human   = suppressWarnings(as.numeric(str_trim(as.character(genomeInsertionPoint)))),
      sr_num      = suppressWarnings(as.numeric(srNum))
    ) %>%
    filter(!is.na(sr_num), sr_num >= min_sr) %>%
    mutate(site_bin = paste(chr_human, floor(pos_human / bin_bp) * bin_bp, sep = ":")) %>%
    group_by(sample_id) %>%
    summarise(
      hpv_type           = first(na.omit(hpv_type)),
      n_breakpoints      = n(),
      n_sites            = n_distinct(site_bin),
      has_reciprocal     = any(orientation == "hv") & any(orientation == "vh"),
      n_distinct_chr     = n_distinct(chr_human),
      chr_coord_range_mb  = if (n_distinct(chr_human) == 1)
                              (max(pos_human, na.rm = TRUE) - min(pos_human, na.rm = TRUE)) / 1e6
                            else NA_real_,
      .groups = "drop"
    ) %>%
    mutate(
      # Non-contiguous human flanks (different chromosomes, or very far apart
      # on the same chromosome) are the chimeric-episome signal.
      flanks_noncontiguous = (n_distinct_chr > 1) |
                             (!is.na(chr_coord_range_mb) &
                                chr_coord_range_mb > params$distant_flank_mb),
      has_hv_junctions     = n_breakpoints > 0
    )
}

shpv_raw <- if (file.exists(params$searchhpv_txt))
  read_tsv(params$searchhpv_txt, show_col_types = FALSE) else NULL
shpv_features <- parse_searchpv(shpv_raw)

cat(sprintf("searcHPV: %d samples with junctions (srNum >= %d).\n",
            nrow(shpv_features), params$min_sr_junc))
head(shpv_features, 8) %>% kable() %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

## 2.2 AmpliconArchitect — architecture features

```{r parse-aa}
# Real columns of interest: "Sample name", "Classification" (Linear/Virus/
# Cyclic/BFB/...), "Location" (e.g. "['chr19:...','hpv16ref_1:0-7909']"),
# "Feature maximum copy number".

parse_aa <- function(df, focal_cn = params$focal_cn) {
  if (is.null(df) || nrow(df) == 0) {
    return(tibble(sample_id = character(), aa_has_hpv = logical(),
                  aa_hpv_cyclic = logical(), aa_cyclic_has_human = logical(),
                  aa_max_cn = double(), aa_focal_amp = logical()))
  }

  df %>%
    transmute(
      sample_id     = `Sample name`,
      classification = Classification,
      location       = Location,
      max_cn         = suppressWarnings(as.numeric(`Feature maximum copy number`)),
      has_hpv_loc    = str_detect(tolower(coalesce(Location, "")), "hpv"),
      has_human_loc  = str_detect(coalesce(Location, ""), "chr"),
      is_cyclic      = str_detect(tolower(coalesce(Classification, "")), "cyclic"),
      is_virus       = str_detect(tolower(coalesce(Classification, "")), "virus")
    ) %>%
    group_by(sample_id) %>%
    summarise(
      aa_has_hpv          = any(has_hpv_loc | is_virus),
      # HPV present inside a cyclic (circular) amplicon = episomal / ecDNA signal
      aa_hpv_cyclic       = any(is_cyclic & has_hpv_loc),
      # cyclic amplicon carrying BOTH human and viral sequence = chimeric ecDNA
      aa_cyclic_has_human = any(is_cyclic & has_hpv_loc & has_human_loc),
      aa_max_cn           = suppressWarnings(max(max_cn, na.rm = TRUE)),
      .groups = "drop"
    ) %>%
    mutate(
      aa_max_cn    = ifelse(is.finite(aa_max_cn), aa_max_cn, NA_real_),
      aa_focal_amp = !is.na(aa_max_cn) & aa_max_cn >= focal_cn
    )
}

aa_raw <- if (file.exists(params$aa_tsv))
  read_tsv(params$aa_tsv, show_col_types = FALSE) else NULL
aa_features <- parse_aa(aa_raw)

cat(sprintf("AmpliconArchitect: %d samples.\n", nrow(aa_features)))
head(aa_features, 8) %>% kable() %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

## 2.3 Coverage z-scores — profile features

```{r parse-cov}
# Real layout: column 1 = "file.name" (sample), remaining columns = per-base
# positions (1,2,3,...) holding the z-score of coverage vs the cohort average.

parse_cov <- function(df, z = params$z_cut,
                      uniform_frac = params$uniform_frac,
                      excess_frac  = params$excess_gain_frac) {
  if (is.null(df) || nrow(df) == 0) {
    return(tibble(sample_id = character(), cov_frac_gain = double(),
                  cov_frac_dip = double(), cov_mean_z = double(),
                  cov_uniform = logical(), cov_n_gain_seg = integer(),
                  cov_excess = logical()))
  }

  id_col <- names(df)[1]                       # "file.name"
  ids    <- df[[id_col]]
  mat    <- as.matrix(df[, setdiff(names(df), id_col)])
  storage.mode(mat) <- "double"

  frac_gain <- rowMeans(mat >= z,  na.rm = TRUE)
  frac_dip  <- rowMeans(mat <= -z, na.rm = TRUE)
  mean_z    <- rowMeans(mat,        na.rm = TRUE)

  # count separated runs of contiguous gain = candidate integration instances
  n_gain_seg <- apply(mat, 1, function(v) {
    g <- !is.na(v) & v >= z
    if (!any(g)) return(0L)
    r <- rle(g); sum(r$values)
  })

  tibble(
    sample_id      = ids,
    cov_frac_gain  = frac_gain,
    cov_frac_dip   = frac_dip,
    cov_mean_z     = mean_z,
    cov_uniform    = (frac_gain + frac_dip) < uniform_frac,
    cov_n_gain_seg = as.integer(n_gain_seg),
    cov_excess     = frac_gain >= excess_frac
  )
}

cov_raw <- if (file.exists(params$zscore_csv))
  read_csv(params$zscore_csv, show_col_types = FALSE) else NULL
cov_features <- parse_cov(cov_raw)

cat(sprintf("Coverage: %d samples profiled.\n", nrow(cov_features)))
head(cov_features, 8) %>% kable(digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

## 2.4 Unified feature table

The three feature tables are joined on `sample_id`. The real input files come
from three different cohorts (their sample naming does not overlap), so on the
real data each row typically carries only one data type. That is expected; the
join is the mechanism, and Section 5 supplies an aligned test set where all
three line up per sample.

```{r unify}
build_unified <- function(shpv, aa, cov) {
  ids <- unique(c(shpv$sample_id, aa$sample_id, cov$sample_id))
  tibble(sample_id = ids) %>%
    left_join(shpv, by = "sample_id") %>%
    left_join(aa,   by = "sample_id") %>%
    left_join(cov,  by = "sample_id") %>%
    mutate(
      has_hv_junctions = coalesce(has_hv_junctions, FALSE),
      n_breakpoints    = coalesce(n_breakpoints, 0L),
      n_sites          = coalesce(n_sites, 0L)
    )
}

unified_real <- build_unified(shpv_features, aa_features, cov_features)
cat(sprintf("Unified real table: %d samples (mostly single-data-type rows).\n",
            nrow(unified_real)))
```

------------------------------------------------------------------------

# 3 Decision logic

Each sample is resolved by combining the three evidence streams. The order
follows the classification diagram (junctions first, then architecture, then
coverage as a modifier).

| Step | Question | Evidence used |
|------|----------|---------------|
| Q1 | Human-viral junctions present? | searcHPV `has_hv_junctions` |
| Q2 | (no junctions) Coverage uniform? | z-score profile `cov_uniform` |
| Q3 | (junctions) Chimeric episome? | AA cyclic-with-human OR non-contiguous human flanks |
| Q4 | Single vs multiple integration sites? | searcHPV `n_sites` (cross-checked by coverage gain segments) |
| Q5 | Episomal modifier? | AA HPV-cyclic OR broad excess coverage gain |

Coverage reading inside this logic: with **no junctions**, a uniform profile is
episomal; internal dips/peaks within the viral genome (non-uniform, no human
sequence) indicate **rearrangement**. With **junctions present**, broad excess
gain beyond what the junctions explain, or a circular HPV amplicon, flags a
coexisting **episomal pool**.

> Subclonal integration is intentionally **not** called here. Its definition
> depends on junction VAF versus tumour purity, and purity is outside these
> three data types, so it is deferred.

```{r classify-fn}
classify_state <- function(r) {

  hv              <- isTRUE(r$has_hv_junctions)
  cyclic_hpv      <- isTRUE(r$aa_hpv_cyclic)
  cyclic_human    <- isTRUE(r$aa_cyclic_has_human)
  noncontig_flank <- isTRUE(r$flanks_noncontiguous)
  uniform         <- isTRUE(r$cov_uniform)
  excess          <- isTRUE(r$cov_excess)
  n_sites         <- ifelse(is.na(r$n_sites), 0L, r$n_sites)
  has_cov         <- !is.na(r$cov_frac_gain)

  # ── Q1: no human-viral junctions → episomal family ──────────────────────
  if (!hv) {
    if (!r$aa_has_hpv %in% TRUE && nrow_cov_missing(r)) return("Unclassified")
    # Q2: uniform coverage = intact episome; non-uniform within HPV = rearranged
    if (has_cov && !uniform) return("HPV episomal rearranged")
    return("Episomal only")
  }

  # ── Junctions present ───────────────────────────────────────────────────
  # Q3: chimeric episome — viral+human on a CIRCULAR molecule. A chimeric
  #     episome is ecDNA (AA cyclic) carrying human + viral sequence. Multi-site
  #     integration also produces junctions on multiple chromosomes, so
  #     non-contiguous flanks only count as chimeric when the molecule is also
  #     circular (otherwise they are just separate linear integration sites).
  if (cyclic_human || (noncontig_flank && cyclic_hpv))
    return("Human-HPV chimeric episome")

  # Q5: episomal modifier — circular HPV amplicon, or coverage far in excess of
  #     what the junctions alone would produce (a free episomal pool).
  if (cyclic_hpv || excess) return("Integrated + episomal (mixed)")

  # Q4: clonal integration — single vs multiple sites. Coverage gain segments
  #     are used as a tie-breaker when junction site count is ambiguous.
  if (n_sites >= 2) return("Integrated clonal, multiple sites")
  if (n_sites == 1) {
    if (!is.na(r$cov_n_gain_seg) && r$cov_n_gain_seg >= 2)
      return("Integrated clonal, multiple sites")
    return("Integrated clonal, single site")
  }
  "Unclassified"
}

# helper: TRUE when this sample has no coverage row at all
nrow_cov_missing <- function(r) is.na(r$cov_frac_gain)
```

------------------------------------------------------------------------

# 4 Apply the classifier

```{r apply-fn}
classify_table <- function(unified) {
  if (nrow(unified) == 0)
    return(mutate(unified, state = factor(character(), levels = state_levels)))
  states <- pmap_chr(unified, function(...) classify_state(tibble(...)))
  mutate(unified, state = factor(states, levels = state_levels))
}
```

------------------------------------------------------------------------

# 5 Synthetic aligned test set

Because the real files are different cohorts, here is a small set of six
engineered samples where all three data types line up per sample, each built to
land in a known state. The same parsers and classifier from above are reused,
so this also exercises the loaders end to end.

```{r synth-data}
# ── Synthetic searcHPV rows ────────────────────────────────────────────────
synth_shpv <- tribble(
  ~Sample,                ~orientation, ~genomeInsertionChromosome, ~genomeInsertionPoint, ~hpvInsertionPoint, ~SiteConfidence,        ~ContigConfidence,        ~srNum,
  # S_single: one reciprocal pair, same locus
  "S_single__HPV16",      "hv",         "chr13",                    34110000,              1200,               "High Confidence Site", "High Confidence Contig", 120,
  "S_single__HPV16",      "vh",         "chr13",                    34110150,              6500,               "High Confidence Site", "High Confidence Contig", 110,
  # S_multi: two distinct loci on different chromosomes
  "S_multi__HPV16",       "hv",         "chr8",                     128700000,             900,                "High Confidence Site", "High Confidence Contig", 90,
  "S_multi__HPV16",       "vh",         "chr8",                     128700120,             5200,               "High Confidence Site", "High Confidence Contig", 85,
  "S_multi__HPV16",       "hv",         "chr3",                     61500000,              3100,               "High Confidence Site", "High Confidence Contig", 80,
  "S_multi__HPV16",       "vh",         "chr3",                     61500140,              7100,               "High Confidence Site", "High Confidence Contig", 78,
  # S_chim: junctions whose human flanks are on different chromosomes
  "S_chim__HPV16",        "hv",         "chr7",                     55000000,              1000,               "High Confidence Site", "High Confidence Contig", 60,
  "S_chim__HPV16",        "vh",         "chr12",                    25000000,              6000,               "High Confidence Site", "High Confidence Contig", 58,
  # S_mixed: one integration locus (episomal pool comes from AA + coverage)
  "S_mixed__HPV16",       "hv",         "chr5",                     1295000,               1500,               "High Confidence Site", "High Confidence Contig", 70,
  "S_mixed__HPV16",       "vh",         "chr5",                     1295160,               6800,               "High Confidence Site", "High Confidence Contig", 66
  # S_epi and S_rearr deliberately have NO human-viral junctions
)

# ── Synthetic AA rows ──────────────────────────────────────────────────────
synth_aa <- tribble(
  ~`Sample name`, ~Classification, ~Location,                                   ~`Feature maximum copy number`,
  "S_epi",        "Virus",         "['hpv16ref_1:0-7909']",                     8,
  "S_rearr",      "Virus",         "['hpv16ref_1:0-7909']",                     12,
  "S_chim",       "Cyclic",        "['chr7:55000000-55050000','hpv16ref_1:0-7909','chr12:25000000-25040000']", 45,
  "S_single",     "Linear",        "['chr13:34100000-34120000','hpv16ref_1:0-7909']", 6,
  "S_multi",      "Linear",        "['chr8:128690000-128710000','hpv16ref_1:0-7909']", 5,
  "S_multi",      "Linear",        "['chr3:61490000-61510000']",                4,
  "S_mixed",      "Linear",        "['chr5:1290000-1300000','hpv16ref_1:0-7909']", 7,
  "S_mixed",      "Cyclic",        "['hpv16ref_1:0-7909']",                     30
)

# ── Synthetic coverage z-score matrix (200 bases) ──────────────────────────
set.seed(1)
n_pos <- 200
mk_profile <- function(kind) {
  z <- rnorm(n_pos, 0, 0.3)                       # baseline noise
  if (kind == "uniform")      z <- z
  if (kind == "internal_dip") z[60:100] <- z[60:100] - 4      # dip inside HPV
  if (kind == "broad_gain")   z <- z + 3                       # whole-genome amplified
  if (kind == "one_gain")     z[80:110] <- z[80:110] + 4       # one gained block
  if (kind == "two_gains")  { z[30:55]  <- z[30:55]  + 4
                              z[130:160] <- z[130:160] + 4 }    # two gained blocks
  if (kind == "excess")     { z <- z + 2; z[90:120] <- z[90:120] + 3 }  # baseline lift + bump
  z
}
synth_cov <- bind_rows(
  c(list(file.name = "S_epi"),    setNames(as.list(mk_profile("uniform")),      1:n_pos)),
  c(list(file.name = "S_rearr"),  setNames(as.list(mk_profile("internal_dip")), 1:n_pos)),
  c(list(file.name = "S_chim"),   setNames(as.list(mk_profile("broad_gain")),   1:n_pos)),
  c(list(file.name = "S_single"), setNames(as.list(mk_profile("one_gain")),     1:n_pos)),
  c(list(file.name = "S_multi"),  setNames(as.list(mk_profile("two_gains")),    1:n_pos)),
  c(list(file.name = "S_mixed"),  setNames(as.list(mk_profile("excess")),       1:n_pos))
)
```

```{r synth-classify}
unified_synth <- build_unified(
  parse_searchpv(synth_shpv),
  parse_aa(synth_aa),
  parse_cov(synth_cov)
)

results_synth <- classify_table(unified_synth)
```

------------------------------------------------------------------------

# 6 Final classification

The deliverable: one row per sample with its assigned HPV state and the
evidence that drove it.

```{r final-table}
final_df <- results_synth %>%
  transmute(
    sample_id,
    state,
    hpv_type,
    junctions      = has_hv_junctions,
    n_sites,
    aa_hpv_cyclic,
    aa_cyclic_has_human,
    flanks_noncontiguous,
    cov_uniform,
    cov_n_gain_seg,
    cov_excess
  ) %>%
  arrange(state)

write_tsv(final_df, file.path(params$out_dir, "hpv_state_classification.tsv"))

final_df %>%
  kable(caption = "HPV state classification (synthetic aligned test set)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE)
```

```{r final-summary}
final_df %>%
  count(state, name = "n", .drop = FALSE) %>%
  ggplot(aes(n, fct_rev(state), fill = state)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = n), hjust = -0.4, size = 4) +
  scale_fill_manual(values = state_palette) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Samples per HPV state", x = "Samples", y = NULL) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.y = element_blank())
```

The same `classify_table()` runs on the real unified table; rows there carry
only partial evidence (different cohorts), so most resolve to `Episomal only`
or `Unclassified` until the three assays are run on shared samples.

```{r real-classify}
results_real <- classify_table(unified_real)
results_real %>% count(state, name = "n", .drop = FALSE) %>%
  kable(caption = "Real data (partial evidence per cohort)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

------------------------------------------------------------------------

# 7 Coordinate concordance (where the three agree)

The classifier above works on per-sample summaries. This section is the
positional analysis you asked for: it places all three data types on a shared
coordinate axis and finds regions where a **coverage segment edge**, a
**searcHPV junction**, and an **AmpliconArchitect boundary** coincide within a
window (`params$overlap_window_bp`, default `r params$overlap_window_bp` bp).

A coverage breakpoint here is the **edge of a sustained segment**: a run of
bases continuously elevated (or depressed) that averages beyond |z| >= 1.5 and
then steps back to normal. Brief sub-threshold flickers inside a run are bridged
(`cov_gap_tol`) and short noise spikes are dropped (`cov_min_seg_len`), so a
single integrated viral block yields exactly two breakpoints (its start and
end), not a scatter of crossings.

**Concordance is expected to be partial.** The three assays will often disagree
on a given edge, by design: overlapping integration segments merge in coverage,
AA may not resolve a sub-range, a junction may have no coverage shift. Every
region is therefore reported with `n_types` (1, 2, or 3) and a `support` label,
and nothing is dropped for lacking all three. A region with all three is the
strongest call; two-of-three is still informative.

The axis matters. Coverage z-scores are indexed along the **HPV genome**, and
searcHPV gives a viral coordinate (`hpvInsertionPoint`) for every junction, so
the HPV axis is the only place all three data types can be compared directly.
AA viral intervals (`hpv16ref_1:start-end`) supply the third anchor there. A
separate **human-axis** check asks whether junction human breakpoints fall
inside AA human amplicon intervals (coverage cannot participate, since the
z-profile is viral-only).

Overlap is reported, not required. Each region is tagged with which of the
three types support it (`n_types` = 1, 2, or 3), so partial agreement is
visible rather than discarded.

## 7.1 Concordance functions

```{r concord-fns}
# A coverage "transition" is the EDGE of a sustained segment: a run of bases
# that is continuously elevated (or depressed) and averages beyond |z| >= z_cut,
# then steps back to normal. We detect the segments, then the breakpoint anchors
# are the segment start/end positions.
#
# gap_tol lets brief dips below threshold stay inside one segment; min_len drops
# short noise spikes; the mean-over-segment test enforces "continuously over 1.5
# on average".
seg_from_flag <- function(flag, gap_tol) {
  idx <- which(flag)
  if (length(idx) == 0) return(list())
  brk <- c(0L, which(diff(idx) > (gap_tol + 1L)), length(idx))
  lapply(seq_len(length(brk) - 1L), function(j) {
    grp <- idx[(brk[j] + 1L):brk[j + 1L]]
    c(min(grp), max(grp))
  })
}

cov_segments <- function(zvec, cut = params$z_cut,
                         gap_tol = params$cov_gap_tol,
                         min_len = params$cov_min_seg_len) {
  out <- list()
  for (dir in c(1L, -1L)) {
    flag <- if (dir == 1L) zvec >= cut else zvec <= -cut
    for (m in seg_from_flag(flag, gap_tol)) {
      s <- m[1]; e <- m[2]
      if (e - s + 1L < min_len) next
      mz <- mean(zvec[s:e])
      if ((dir == 1L && mz >= cut) || (dir == -1L && mz <= -cut)) {
        out[[length(out) + 1L]] <- tibble(seg_start = s, seg_end = e,
                                          type = if (dir == 1L) "gain" else "dip",
                                          mean_z = mz, len = e - s + 1L)
      }
    }
  }
  if (length(out) == 0)
    return(tibble(seg_start = integer(), seg_end = integer(),
                  type = character(), mean_z = double(), len = integer()))
  bind_rows(out) %>% arrange(seg_start)
}

# Breakpoint anchors contributed by coverage = the edges of each sustained segment.
cov_edges <- function(zvec, ...) {
  segs <- cov_segments(zvec, ...)
  if (nrow(segs) == 0) return(integer())
  sort(unique(c(segs$seg_start, segs$seg_end)))
}

# searcHPV viral coordinate of each junction for one sample.
junction_viral_pos <- function(df, sid) {
  df %>%
    mutate(sample_id = str_remove(Sample, "__HPV\\d+.*$"),
           hpv_pos   = suppressWarnings(as.numeric(str_trim(as.character(hpvInsertionPoint))))) %>%
    filter(sample_id == sid, !is.na(hpv_pos)) %>% pull(hpv_pos)
}

# AA viral interval boundaries (start and end of every hpv interval) for one sample.
aa_viral_boundaries <- function(df, sid) {
  rows <- df %>% filter(`Sample name` == sid)
  if (nrow(rows) == 0) return(numeric(0))
  locs <- unlist(str_extract_all(rows$Location, "hpv[^']*?:\\d+-\\d+"))
  if (length(locs) == 0) return(numeric(0))
  m <- str_match(locs, ":(\\d+)-(\\d+)")
  as.numeric(c(m[, 2], m[, 3]))
}

# Cluster all anchors within W bp and tag which data types support each region.
build_concordance <- function(sid, zvec, shpv_df, aa_df,
                              W = params$overlap_window_bp) {
  anchors <- bind_rows(
    tibble(source = "coverage", pos = cov_edges(zvec)),
    tibble(source = "junction", pos = junction_viral_pos(shpv_df, sid)),
    tibble(source = "AA",       pos = aa_viral_boundaries(aa_df, sid))
  ) %>% filter(!is.na(pos)) %>% arrange(pos)
  if (nrow(anchors) == 0) return(tibble())

  grp <- integer(nrow(anchors)); g <- 0L; last <- -Inf
  for (i in seq_len(nrow(anchors))) {
    if (anchors$pos[i] - last > W) g <- g + 1L
    grp[i] <- g; last <- anchors$pos[i]
  }
  anchors$region <- grp

  anchors %>%
    group_by(region) %>%
    summarise(
      sample_id = sid,
      hpv_start = min(pos), hpv_end = max(pos),
      coverage  = any(source == "coverage"),
      junction  = any(source == "junction"),
      AA        = any(source == "AA"),
      .groups   = "drop"
    ) %>%
    mutate(n_types = coverage + junction + AA,
           support = case_when(n_types == 3 ~ "all three",
                               n_types == 2 ~ "two of three",
                               TRUE         ~ "single")) %>%
    select(sample_id, region, hpv_start, hpv_end,
           coverage, junction, AA, n_types, support)
}

# Human-axis supplement: junction human breakpoint inside an AA human interval.
aa_human_intervals <- function(df, sid) {
  rows <- df %>% filter(`Sample name` == sid)
  locs <- unlist(str_extract_all(rows$Location, "chr[^']*?:\\d+-\\d+"))
  if (length(locs) == 0) return(tibble(chr = character(), start = double(), end = double()))
  m <- str_match(locs, "(chr[\\dXYM]+):(\\d+)-(\\d+)")
  tibble(chr = m[, 2], start = as.numeric(m[, 3]), end = as.numeric(m[, 4]))
}

human_overlap <- function(shpv_df, aa_df, sid) {
  jh <- shpv_df %>%
    mutate(sample_id = str_remove(Sample, "__HPV\\d+.*$")) %>%
    filter(sample_id == sid) %>%
    transmute(chr = genomeInsertionChromosome,
              pos = suppressWarnings(as.numeric(str_trim(as.character(genomeInsertionPoint)))))
  ai <- aa_human_intervals(aa_df, sid)
  if (nrow(jh) == 0) return(tibble())
  jh %>%
    rowwise() %>%
    mutate(in_aa_interval = any(ai$chr == chr & ai$start <= pos & ai$end >= pos)) %>%
    ungroup() %>%
    mutate(sample_id = sid) %>%
    select(sample_id, chr, pos, in_aa_interval)
}
```

## 7.2 Aligned demonstration

Two integration samples on a full-length viral axis (HPV16, ~7906 bp). The
coverage gain blocks span each integrated viral segment, and the AA viral
intervals are the matching sub-ranges, so segment edges, junctions, and AA
boundaries should line up. `S_multi` has two integration sites whose viral
segments partly overlap: in coverage they merge into one sustained block, so
coverage contributes only the outer two edges while junctions and AA still mark
all four. That is exactly the partial-concordance case, the inner edges resolve
to two-of-three rather than all three.

```{r concord-demo}
hpv_len <- 7906
mk_full <- function(blocks, len = hpv_len, sd = 0.3) {
  z <- rnorm(len, 0, sd)
  for (b in blocks) z[b[1]:b[2]] <- z[b[1]:b[2]] + b[3]
  z
}
set.seed(2)
zvecs <- list(
  S_single = mk_full(list(c(1200, 6500, 4))),
  S_multi  = mk_full(list(c(900, 5200, 4), c(3100, 7100, 4)))
)

# AA frame for the demo: viral sub-ranges matching the integrated segments
# (kept separate from the classifier's synth_aa so Sections 5-6 are untouched).
aa_demo <- tribble(
  ~`Sample name`, ~Classification, ~Location,                                          ~`Feature maximum copy number`,
  "S_single",     "Linear",        "['chr13:34100000-34120000','hpv16ref_1:1200-6500']", 6,
  "S_multi",      "Linear",        "['chr8:128690000-128710000','hpv16ref_1:900-5200']", 5,
  "S_multi",      "Linear",        "['chr3:61490000-61510000','hpv16ref_1:3100-7100']",  4
)

concordance_tbl <- bind_rows(
  build_concordance("S_single", zvecs$S_single, synth_shpv, aa_demo),
  build_concordance("S_multi",  zvecs$S_multi,  synth_shpv, aa_demo)
)

write_tsv(concordance_tbl, file.path(params$out_dir, "hpv_axis_concordance.tsv"))

concordance_tbl %>%
  kable(caption = "HPV-axis overlap regions (coverage transition / junction / AA boundary)",
        col.names = c("Sample", "Region", "HPV start", "HPV end",
                      "Coverage", "Junction", "AA", "N types", "Support")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) %>%
  row_spec(which(concordance_tbl$n_types == 3), background = "#e6f4ea")
```

## 7.3 Human-axis check

```{r concord-human}
human_tbl <- bind_rows(
  human_overlap(synth_shpv, aa_demo, "S_single"),
  human_overlap(synth_shpv, aa_demo, "S_multi")
)

human_tbl %>%
  kable(caption = "Junction human breakpoints vs AA human amplicon intervals",
        col.names = c("Sample", "Chr", "Human pos", "Inside AA interval?")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

A region supported by **all three** on the HPV axis, whose human breakpoint
also sits inside an AA human interval, is the strongest possible integration
call from these inputs: the junction, the architecture, and the coverage shift
all point to the same place.

------------------------------------------------------------------------

# 8 Session info

```{r session-info, echo=FALSE}
sessionInfo()
```
