---
title: "HPV State Classification — OPC real data (HN-keyed)"
subtitle: "searcHPV V2 + per-base coverage z-scores, AmpliconArchitect optional"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 3
    code_folding: show
    df_print: paged
params:
  # ── Inputs ────────────────────────────────────────────────────────────────
  summary_xlsx:  "data/summ_OPC.v7.2.xlsx"                 # crosswalk: seqname <-> HN id
  searchhpv_txt: "data/2024-05-27_SearcHPV_OPC_V2.txt"     # OPC integration (V2 schema)
  zscore_csv:    "data/20241205_Zscores_stats_nofilter.csv" # OPC per-base coverage z-scores
  aa_tsv:        "data/merged_result_tables.tsv"            # AA (NOT OPC yet) - only used if use_aa
  # ── The toggle you asked for ──────────────────────────────────────────────
  use_aa:             false   # FALSE = run on searcHPV + coverage only (current OPC state)
  # ── Timepoint filter ──────────────────────────────────────────────────────
  timepoint_filter:   "Baseline"   # keep only this timepoint; set "" or null for all
  # ── Thresholds (tune here) ────────────────────────────────────────────────
  high_conf_site_only: false  # TRUE = keep only "High Confidence Site" searcHPV breakpoints
  site_bin_bp:        200     # window (bp) to collapse nearby breakpoints into one site
  z_cut:              1.5     # |z| >= this = coverage dip (negative) or gain (positive)
  uniform_frac:       0.05    # if < this fraction of bases deviate, coverage is "uniform"
  excess_gain_frac:   0.50    # if >= this fraction of bases gained, treat as episomal pool
  distant_flank_mb:   10      # human flanks farther apart than this (same chr) = chimeric signal
  focal_cn:           10      # AA max copy number above this = focal amplification
  overlap_window_bp:  200     # window (bp) to call a coverage/junction/AA overlap
  cov_gap_tol:        25      # bp: merge elevated runs separated by gaps <= this
  cov_min_seg_len:    30      # bp: discard sustained segments shorter than this (noise)
  n_plot_samples:     6       # how many integrated samples to draw track PDFs for
  out_dir:            "results"   # TSV tables
  fig_dir:            "plots"     # per-sample plot PDFs go in this sub-directory
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6.5)
```

This version runs the **real OPC data** and uses the **HN identifier**
(the summary's `id` = `Patient_ID`\_`Timepoint`, e.g. `HN2186_Baseline`)
as the reference key throughout. The three sources are tied together by
the summary crosswalk:

-   **searcHPV** (`2024-05-27_SearcHPV_OPC_V2.txt`) — integration
    junctions. Its `Sample` is a sequencing-library name; it also
    carries `Patient_ID`, `Timepoint`, and a ready-made `id`.
-   **Coverage** (`20241205_Zscores_stats_nofilter.csv`) — per-base
    z-scores, keyed only by sequencing-library name (`file.name`).
-   **Summary** (`summ_OPC.v7.2.xlsx`) — maps every sequencing-library
    `Sample` to its HN `id`, `Patient_ID`, and `Timepoint`. This is the
    crosswalk that lets coverage be relabelled to HN ids.

> **HN id is per-timepoint.** Because integration and coverage differ
> between a patient's timepoints, the working key is the full `id` (HN +
> timepoint). The bare HN (`Patient_ID`) is also carried as a column so
> you can group across timepoints.

## The AA toggle

`params$use_aa` switches AmpliconArchitect on or off. OPC has no AA yet,
so the default is **`use_aa: false`** and the run uses searcHPV +
coverage only. When you have OPC AA, point `params$aa_tsv` at it and set
`use_aa: true` (you can also flip it in the RStudio *Knit with
Parameters* dialog).

**What AA off costs you.** AA is the only source for circular-molecule
evidence, so with it off the classifier cannot positively call
`Human-HPV chimeric episome` or the AA-driven
`Integrated + episomal (mixed)`. Those samples instead resolve to the
best junction/coverage call (usually multi-site integration), and the
`notes` column records that AA was unavailable. Everything else
(episomal vs rearranged, single vs multiple integration sites) is
unaffected.

------------------------------------------------------------------------

# 1 Packages and labels

```{r packages}
suppressPackageStartupMessages({
  library(tidyverse)
  library(readxl)
  library(knitr)
  library(kableExtra)
})

dir.create(params$out_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(params$fig_dir, showWarnings = FALSE, recursive = TRUE)

state_levels <- c(
  "Episomal only",
  "HPV episomal rearranged",
  "Human-HPV chimeric episome",
  "Integrated clonal, single site",
  "Integrated clonal, multiple sites",
  "Integrated + episomal (mixed)",
  "Unclassified"
)
state_palette <- c(
  "Episomal only"                     = "#1D9E75",
  "HPV episomal rearranged"           = "#3EC9A7",
  "Human-HPV chimeric episome"        = "#D4537E",
  "Integrated clonal, single site"    = "#534AB7",
  "Integrated clonal, multiple sites" = "#3A2F8F",
  "Integrated + episomal (mixed)"     = "#BA7517",
  "Unclassified"                      = "#888780"
)

cat(sprintf("AmpliconArchitect: %s\n", if (isTRUE(params$use_aa)) "ENABLED" else "DISABLED (searcHPV + coverage only)"))
```

------------------------------------------------------------------------

# 2 Crosswalk (HN id is the reference)

```{r crosswalk}
# NOTE: the sheet has TWO columns named "Sample" (col 1 = full sequencing name
# with _L00x, col 2 = without). readxl renames duplicates, so we grab the first
# by position - that is the one matching coverage file.name and searcHPV Sample.
summ_raw <- suppressMessages(read_excel(params$summary_xlsx, .name_repair = "unique"))
seqname_col <- names(summ_raw)[1]

summ <- summ_raw %>%
  transmute(seqname   = as.character(.data[[seqname_col]]),
            sample_id = as.character(id),          # HN_Timepoint = the reference key
            patient   = as.character(Patient_ID),  # bare HN
            timepoint = as.character(Timepoint),
            hpv_type  = as.character(HPV_realign))

# Timepoint filter (default Baseline). Cascades to coverage (via the crosswalk)
# and to searcHPV (filtered in the load chunk).
tp_filter <- params$timepoint_filter
if (!is.null(tp_filter) && nzchar(tp_filter)) {
  summ <- filter(summ, timepoint == tp_filter)
  cat(sprintf("Timepoint filter: keeping only '%s'.\n", tp_filter))
} else {
  cat("Timepoint filter: none (all timepoints).\n")
}

# seqname -> HN id, and the reverse (id is unique per seqname)
seq2id <- setNames(summ$sample_id, summ$seqname)
id2seq <- setNames(summ$seqname,   summ$sample_id)

meta <- summ %>% distinct(sample_id, patient, timepoint, hpv_type)

cat(sprintf("Crosswalk: %d sequencing libraries -> %d HN ids.\n",
            nrow(summ), n_distinct(summ$sample_id)))
head(summ, 5) %>% kable() %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

------------------------------------------------------------------------

# 3 Parsers

Each parser returns a per-sample feature table keyed by the HN
`sample_id`.

## 3.1 searcHPV (V2 schema)

The V2 file has no `orientation` or `srNum` columns, so reciprocal-pair
and read-count filters from earlier versions are dropped. Breakpoints
are filtered by `SiteConfidence` instead (optional, via
`high_conf_site_only`), and the HN id is taken from the file's own `id`
column when present, otherwise resolved through the crosswalk, otherwise
derived from the sample name (so synthetic tests still work).

```{r parse-searchpv}
parse_searchpv <- function(df,
                           high_conf_site_only = params$high_conf_site_only,
                           bin_bp = params$site_bin_bp) {
  empty <- tibble(sample_id = character(), hpv_type = character(),
                  n_breakpoints = integer(), n_sites = integer(),
                  n_distinct_chr = integer(), chr_coord_range_mb = double(),
                  flanks_noncontiguous = logical(), has_hv_junctions = logical())
  if (is.null(df) || nrow(df) == 0) return(empty)

  sid <- if ("id" %in% names(df)) as.character(df$id) else
           ifelse(as.character(df$Sample) %in% names(seq2id),
                  unname(seq2id[as.character(df$Sample)]),
                  str_remove(as.character(df$Sample), "__HPV\\d+.*$"))
  hpv <- if ("HPV_realign" %in% names(df)) as.character(df$HPV_realign) else
           str_extract(as.character(df$Sample), "HPV\\d+")
  conf <- if ("SiteConfidence" %in% names(df)) as.character(df$SiteConfidence) else NA_character_

  work <- tibble(
    sample_id = sid,
    hpv_type  = hpv,
    chr       = as.character(df$genomeInsertionChromosome),
    pos       = suppressWarnings(as.numeric(str_trim(as.character(df$genomeInsertionPoint)))),
    site_conf = conf
  )
  if (isTRUE(high_conf_site_only) && any(!is.na(work$site_conf)))
    work <- filter(work, str_detect(tolower(coalesce(site_conf, "")), "high"))
  if (nrow(work) == 0) return(empty)

  work %>%
    mutate(site_bin = paste(chr, floor(pos / bin_bp) * bin_bp, sep = ":")) %>%
    group_by(sample_id) %>%
    summarise(
      hpv_type           = first(na.omit(hpv_type)),
      n_breakpoints      = n(),
      n_sites            = n_distinct(site_bin),
      n_distinct_chr     = n_distinct(chr),
      chr_coord_range_mb = if (n_distinct(chr) == 1)
                             (max(pos, na.rm = TRUE) - min(pos, na.rm = TRUE)) / 1e6
                           else NA_real_,
      .groups = "drop"
    ) %>%
    mutate(
      flanks_noncontiguous = (n_distinct_chr > 1) |
                             (!is.na(chr_coord_range_mb) &
                                chr_coord_range_mb > params$distant_flank_mb),
      has_hv_junctions = n_breakpoints > 0
    )
}
```

## 3.2 Coverage z-scores

```{r parse-cov}
parse_cov <- function(df, z = params$z_cut,
                      uniform_frac = params$uniform_frac,
                      excess_frac  = params$excess_gain_frac) {
  empty <- tibble(seqname = character(), cov_frac_gain = double(),
                  cov_frac_dip = double(), cov_mean_z = double(),
                  cov_uniform = logical(), cov_n_gain_seg = integer(),
                  cov_n_dip_seg = integer(), cov_n_event_seg = integer(),
                  cov_excess = logical())
  if (is.null(df) || nrow(df) == 0) return(empty)

  id_col <- names(df)[1]                       # "file.name"
  ids    <- as.character(df[[id_col]])
  mat    <- as.matrix(df[, setdiff(names(df), id_col)])
  storage.mode(mat) <- "double"

  frac_gain <- rowMeans(mat >= z,  na.rm = TRUE)
  frac_dip  <- rowMeans(mat <= -z, na.rm = TRUE)
  mean_z    <- rowMeans(mat,        na.rm = TRUE)

  count_seg <- function(v, up) {
    flag <- if (up) (!is.na(v) & v >= z) else (!is.na(v) & v <= -z)
    if (!any(flag)) return(0L)
    sum(rle(flag)$values)
  }
  n_gain_seg <- apply(mat, 1, count_seg, up = TRUE)
  n_dip_seg  <- apply(mat, 1, count_seg, up = FALSE)

  tibble(
    seqname         = ids,
    cov_frac_gain   = frac_gain,
    cov_frac_dip    = frac_dip,
    cov_mean_z      = mean_z,
    cov_uniform     = (frac_gain + frac_dip) < uniform_frac,
    cov_n_gain_seg  = as.integer(n_gain_seg),
    cov_n_dip_seg   = as.integer(n_dip_seg),
    cov_n_event_seg = as.integer(n_gain_seg + n_dip_seg),
    cov_excess      = frac_gain >= excess_frac
  )
}
```

## 3.3 AmpliconArchitect (only read when `use_aa`)

```{r parse-aa}
parse_aa <- function(df, focal_cn = params$focal_cn) {
  empty <- tibble(sample_id = character(), aa_has_hpv = logical(),
                  aa_hpv_cyclic = logical(), aa_cyclic_has_human = logical(),
                  aa_max_cn = double(), aa_focal_amp = logical())
  if (is.null(df) || nrow(df) == 0) return(empty)

  sid <- as.character(df$`Sample name`)
  sid <- ifelse(sid %in% names(seq2id), unname(seq2id[sid]), sid)  # map to HN id if possible

  df %>%
    mutate(.sid = sid) %>%
    transmute(
      sample_id     = .sid,
      has_hpv_loc   = str_detect(tolower(coalesce(Location, "")), "hpv"),
      has_human_loc = str_detect(coalesce(Location, ""), "chr"),
      is_cyclic     = str_detect(tolower(coalesce(Classification, "")), "cyclic"),
      is_virus      = str_detect(tolower(coalesce(Classification, "")), "virus"),
      max_cn        = suppressWarnings(as.numeric(`Feature maximum copy number`))
    ) %>%
    group_by(sample_id) %>%
    summarise(
      aa_has_hpv          = any(has_hpv_loc | is_virus),
      aa_hpv_cyclic       = any(is_cyclic & has_hpv_loc),
      aa_cyclic_has_human = any(is_cyclic & has_hpv_loc & has_human_loc),
      aa_max_cn           = suppressWarnings(max(max_cn, na.rm = TRUE)),
      .groups = "drop"
    ) %>%
    mutate(aa_max_cn = ifelse(is.finite(aa_max_cn), aa_max_cn, NA_real_),
           aa_focal_amp = !is.na(aa_max_cn) & aa_max_cn >= focal_cn)
}
```

------------------------------------------------------------------------

# 4 Load real data

```{r load}
# searcHPV (apply the same timepoint filter when the column is present)
shpv_raw <- read_tsv(params$searchhpv_txt, show_col_types = FALSE)
if (!is.null(tp_filter) && nzchar(tp_filter) && "Timepoint" %in% names(shpv_raw))
  shpv_raw <- filter(shpv_raw, as.character(Timepoint) == tp_filter)
shpv_features <- parse_searchpv(shpv_raw)

# coverage (large file; read once, keep raw for plotting)
cov_raw       <- read_csv(params$zscore_csv, show_col_types = FALSE)
cov_features  <- parse_cov(cov_raw) %>%
  mutate(sample_id = unname(seq2id[seqname])) %>%   # seqname -> HN id
  filter(!is.na(sample_id)) %>%
  select(sample_id, everything(), -seqname)

# AA only if toggled on
aa_raw <- if (isTRUE(params$use_aa) && file.exists(params$aa_tsv))
  read_tsv(params$aa_tsv, show_col_types = FALSE) else NULL
aa_features <- if (isTRUE(params$use_aa)) parse_aa(aa_raw) else parse_aa(NULL)

cat(sprintf("searcHPV: %d integrated samples\ncoverage: %d samples (HN-mapped)\nAA: %s\n",
            nrow(shpv_features), nrow(cov_features),
            if (isTRUE(params$use_aa)) sprintf("%d samples", nrow(aa_features)) else "disabled"))
```

```{r unify}
build_unified <- function(shpv, aa, cov, meta) {
  ids <- unique(c(shpv$sample_id, aa$sample_id, cov$sample_id))
  tibble(sample_id = ids) %>%
    left_join(shpv, by = "sample_id") %>%
    left_join(aa,   by = "sample_id") %>%
    left_join(cov,  by = "sample_id") %>%
    left_join(meta, by = "sample_id", suffix = c("", "_meta")) %>%
    mutate(
      hpv_type         = coalesce(hpv_type, hpv_type_meta),
      has_hv_junctions = coalesce(has_hv_junctions, FALSE),
      n_breakpoints    = coalesce(n_breakpoints, 0L),
      n_sites          = coalesce(n_sites, 0L)
    ) %>%
    select(-any_of("hpv_type_meta"))
}

unified <- build_unified(shpv_features, aa_features, cov_features, meta)
cat(sprintf("Unified OPC table: %d samples.\n", nrow(unified)))
```

------------------------------------------------------------------------

# 5 Classifier

Site count (single vs multiple) is decided by the searcHPV junction
count. Coverage that exceeds the junction count (extra gains or losses)
is recorded in `notes`, not used to change the site count. AA-dependent
states are only reachable when `use_aa` is on.

```{r classify-fn}
classify_state <- function(r) {
  hv              <- isTRUE(r$has_hv_junctions)
  cyclic_hpv      <- isTRUE(r$aa_hpv_cyclic)
  cyclic_human    <- isTRUE(r$aa_cyclic_has_human)
  noncontig_flank <- isTRUE(r$flanks_noncontiguous)
  uniform         <- isTRUE(r$cov_uniform)
  excess          <- isTRUE(r$cov_excess)
  n_sites         <- ifelse(is.na(r$n_sites), 0L, r$n_sites)
  has_cov         <- !is.na(r$cov_frac_gain)

  if (!hv) {
    if (!isTRUE(r$aa_has_hpv) && !has_cov) return("Unclassified")
    if (has_cov && !uniform) return("HPV episomal rearranged")
    return("Episomal only")
  }
  if (cyclic_human || (noncontig_flank && cyclic_hpv))
    return("Human-HPV chimeric episome")
  if (cyclic_hpv || excess) return("Integrated + episomal (mixed)")
  if (n_sites >= 2) return("Integrated clonal, multiple sites")
  if (n_sites == 1) return("Integrated clonal, single site")
  "Unclassified"
}

add_notes <- function(df, aa_enabled = isTRUE(params$use_aa)) {
  df %>%
    mutate(
      cov_extra = coalesce(has_hv_junctions, FALSE) &
                  coalesce(cov_n_event_seg, 0L) > coalesce(n_sites, 0L),
      no_aa = aa_enabled & is.na(aa_has_hpv),
      notes = pmap_chr(
        list(cov_extra, no_aa, cov_n_event_seg, n_sites),
        function(extra, na_aa, nev, ns) {
          msgs <- character()
          if (isTRUE(extra))
            msgs <- c(msgs, sprintf(
              "coverage shows %d event segment(s) vs %d junction site(s); extra is coverage-only and NOT counted as a site",
              ifelse(is.na(nev), 0L, nev), ifelse(is.na(ns), 0L, ns)))
          if (isTRUE(na_aa))
            msgs <- c(msgs, "no AmpliconArchitect evidence for this sample")
          if (length(msgs) == 0) "" else paste(msgs, collapse = "; ")
        })
    )
}

classify_table <- function(unified, aa_enabled = isTRUE(params$use_aa)) {
  if (nrow(unified) == 0)
    return(mutate(unified, state = factor(character(), levels = state_levels)))
  states <- pmap_chr(unified, function(...) classify_state(tibble(...)))
  unified %>%
    mutate(state = factor(states, levels = state_levels)) %>%
    add_notes(aa_enabled = aa_enabled)
}
```

------------------------------------------------------------------------

# 6 Classify the OPC cohort

```{r classify-run}
results <- classify_table(unified)

state_summary <- results %>% count(state, name = "n", .drop = FALSE) %>% arrange(desc(n))
state_summary %>%
  kable(caption = sprintf("OPC HPV state distribution (use_aa = %s)", params$use_aa)) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)

results %>%
  filter(state != "Unclassified") %>%
  count(state, name = "n", .drop = FALSE) %>%
  ggplot(aes(n, fct_rev(state), fill = state)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = n), hjust = -0.3, size = 4) +
  scale_fill_manual(values = state_palette) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "OPC samples per HPV state", x = "Samples", y = NULL) +
  theme_minimal(base_size = 13) +
  theme(panel.grid.major.y = element_blank())
```

```{r classify-table-out}
out_tbl <- results %>%
  transmute(
    sample_id, patient, timepoint, state, hpv_type,
    junctions = has_hv_junctions, n_sites,
    cov_frac_gain, cov_frac_dip, cov_uniform, cov_excess, cov_n_event_seg,
    aa_hpv_cyclic, aa_cyclic_has_human, notes
  ) %>%
  arrange(patient, timepoint)

write_tsv(out_tbl, file.path(params$out_dir, "hpv_state_OPC.tsv"))

out_tbl %>% head(25) %>%
  kable(caption = "OPC classification (first 25 rows; full table written to results/hpv_state_OPC.tsv)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```

Per-patient view across timepoints (HN as the grouping key):

```{r per-patient}
results %>%
  filter(!is.na(patient)) %>%
  count(patient, state, .drop = FALSE) %>%
  filter(n > 0) %>%
  pivot_wider(names_from = state, values_from = n, values_fill = 0) %>%
  head(20) %>%
  kable(caption = "States per patient across timepoints (first 20 patients)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = TRUE)
```

------------------------------------------------------------------------

# 7 Combined per-sample feature plots

One figure per sample, stacking coverage z-score, searcHPV junctions,
and (when `use_aa`) AA viral intervals on the shared HPV-genome axis.
Each is written as a PDF named `HPVtracks_<HN id>.pdf` to
`params$fig_dir` (default the run directory).

```{r plot-helpers}
seg_from_flag <- function(flag, gap_tol) {
  idx <- which(flag)
  if (length(idx) == 0) return(list())
  brk <- c(0L, which(diff(idx) > (gap_tol + 1L)), length(idx))
  lapply(seq_len(length(brk) - 1L), function(j) {
    grp <- idx[(brk[j] + 1L):brk[j + 1L]]; c(min(grp), max(grp))
  })
}
cov_segments <- function(zvec, cut = params$z_cut,
                         gap_tol = params$cov_gap_tol, min_len = params$cov_min_seg_len) {
  out <- list()
  for (dir in c(1L, -1L)) {
    flag <- if (dir == 1L) (!is.na(zvec) & zvec >= cut) else (!is.na(zvec) & zvec <= -cut)
    for (m in seg_from_flag(flag, gap_tol)) {
      s <- m[1]; e <- m[2]
      if (e - s + 1L < min_len) next
      mz <- mean(zvec[s:e], na.rm = TRUE)
      if ((dir == 1L && mz >= cut) || (dir == -1L && mz <= -cut))
        out[[length(out) + 1L]] <- tibble(seg_start = s, seg_end = e,
                                          type = if (dir == 1L) "gain" else "dip")
    }
  }
  if (length(out) == 0) return(tibble(seg_start = integer(), seg_end = integer(), type = character()))
  bind_rows(out) %>% arrange(seg_start)
}

# Coverage vector for one HN id (NA-trimmed at the tail).
cov_vec <- function(sid) {
  sq <- unname(id2seq[sid])          # single-bracket: returns NA if id absent
  if (length(sq) == 0 || is.na(sq)) return(numeric())
  row <- cov_raw[as.character(cov_raw[[1]]) == sq, , drop = FALSE]
  if (nrow(row) == 0) return(numeric())
  v <- as.numeric(row[1, -1])
  last <- suppressWarnings(max(which(!is.na(v))))
  if (is.finite(last)) v[seq_len(last)] else v
}

# searcHPV viral breakpoint positions for one HN id.
shpv_real <- shpv_raw %>%
  mutate(sample_id = if ("id" %in% names(.)) as.character(id) else
                       ifelse(as.character(Sample) %in% names(seq2id),
                              unname(seq2id[as.character(Sample)]),
                              str_remove(as.character(Sample), "__HPV\\d+.*$")),
         hpv_pos   = suppressWarnings(as.numeric(str_trim(as.character(hpvInsertionPoint)))))
junc_pos <- function(sid) shpv_real %>% filter(sample_id == sid, !is.na(hpv_pos)) %>% pull(hpv_pos)

# AA viral intervals for one HN id (empty when AA off).
aa_intervals <- function(sid) {
  if (!isTRUE(params$use_aa) || is.null(aa_raw)) return(tibble(start = double(), end = double()))
  sn <- names(seq2id)[match(sid, seq2id)]
  rows <- aa_raw %>% filter(as.character(`Sample name`) %in% c(sid, sn))
  if (nrow(rows) == 0) return(tibble(start = double(), end = double()))
  locs <- unlist(str_extract_all(rows$Location, "hpv[^']*?:\\d+-\\d+"))
  if (length(locs) == 0) return(tibble(start = double(), end = double()))
  m <- str_match(locs, ":(\\d+)-(\\d+)")
  tibble(start = as.numeric(m[, 2]), end = as.numeric(m[, 3]))
}
```

```{r plot-fn}
plot_sample_tracks <- function(sid, state_label = "", show_aa = isTRUE(params$use_aa),
                               z_cut = params$z_cut) {
  zvec <- cov_vec(sid)
  jpos <- junc_pos(sid)
  ai   <- aa_intervals(sid)

  trk <- c("Coverage (z-score)", "searcHPV junctions")
  if (show_aa) trk <- c(trk, "AA viral intervals")
  as_trk <- function(x) factor(x, levels = trk)

  cov_df <- if (length(zvec))
    tibble(pos = seq_along(zvec), z = zvec, track = as_trk(trk[1])) else NULL
  segs   <- if (length(zvec)) cov_segments(zvec, cut = z_cut) else tibble()
  seg_df <- if (nrow(segs)) mutate(segs, track = as_trk(trk[1])) else NULL
  thr_df <- tibble(track = as_trk(trk[1]), yint = c(z_cut, -z_cut))
  junc_df <- if (length(jpos)) tibble(pos = jpos, y = 1, track = as_trk(trk[2])) else NULL
  aa_df   <- if (show_aa && nrow(ai)) mutate(ai, y = 1, track = as_trk("AA viral intervals")) else NULL

  ggplot() +
    { if (!is.null(seg_df))
        geom_rect(data = seg_df, aes(xmin = seg_start, xmax = seg_end,
                  ymin = -Inf, ymax = Inf, fill = type), alpha = 0.18, inherit.aes = FALSE) } +
    { if (length(jpos))
        geom_vline(xintercept = jpos, linetype = "dashed", colour = "#534AB7", linewidth = 0.4) } +
    geom_hline(data = thr_df, aes(yintercept = yint), linetype = "dotted", colour = "grey50") +
    { if (!is.null(cov_df)) geom_line(data = cov_df, aes(pos, z), colour = "#222222", linewidth = 0.35) } +
    { if (!is.null(junc_df)) geom_point(data = junc_df, aes(pos, y), colour = "#534AB7", size = 2.6) } +
    { if (!is.null(aa_df)) geom_segment(data = aa_df, aes(x = start, xend = end, y = y, yend = y),
                  colour = "#BA7517", linewidth = 4) } +
    scale_fill_manual(values = c(gain = "#1D9E75", dip = "#D4537E"), name = "Coverage segment") +
    facet_grid(track ~ ., scales = "free_y", switch = "y") +
    labs(title = paste("Feature tracks —", sid),
         subtitle = if (nzchar(state_label)) paste("State:", state_label) else NULL,
         x = "HPV genome position (bp)", y = NULL) +
    theme_minimal(base_size = 12) +
    theme(strip.placement = "outside",
          strip.text.y.left = element_text(angle = 0, face = "bold"),
          panel.grid.minor = element_blank(), legend.position = "bottom")
}
```

```{r choose-and-save}
# integrated samples that also have coverage, most sites first
plot_ids <- results %>%
  filter(has_hv_junctions, !is.na(cov_frac_gain)) %>%
  arrange(desc(n_sites)) %>%
  slice_head(n = params$n_plot_samples) %>%
  transmute(sample_id, state = as.character(state))

saved <- character()
for (i in seq_len(nrow(plot_ids))) {
  sid <- plot_ids$sample_id[i]
  p   <- plot_sample_tracks(sid, state_label = plot_ids$state[i])
  fn  <- file.path(params$fig_dir, paste0("HPVtracks_", gsub("[^A-Za-z0-9_]+", "_", sid), ".pdf"))
  ggsave(fn, p, width = 9, height = if (isTRUE(params$use_aa)) 6.5 else 5)
  saved <- c(saved, normalizePath(fn, mustWork = FALSE))
}
cat("Saved per-sample track PDFs:\n"); cat(paste0("  - ", saved), sep = "\n")
```

```{r show-one, fig.width=9, fig.height=5}
if (nrow(plot_ids)) plot_sample_tracks(plot_ids$sample_id[1], plot_ids$state[1])
```

To plot any specific patient/timepoint, call
`plot_sample_tracks("HN2186_Baseline")`.

------------------------------------------------------------------------

# 8 Session info

```{r session-info, echo=FALSE}
sessionInfo()
```
